# Chapter 3: Hypothesis Testing 

```{r setup}
# Load all libraries needed for this chapter
library(tidyverse)
library(knitr)
library(kableExtra)
library(patchwork)
library(moments)       # skewness
library(coin)          # exact non-parametric tests
library(pwr)           # power analysis
library(effectsize)    # Cohen's d and effect size measures
library(rstatix)       # tidy statistical tests
```

------------------------------------------------------------------------

# Chapter Overview

## Introduction

Hypothesis testing is the formal procedure for using sample data to make
decisions about population parameters. It is the engine behind
scientific claims: *Does this drug reduce blood pressure? Is this
algorithm more accurate than the baseline? Do students in different
programs differ in exam performance?* Every such question can be framed
as a hypothesis test.

This chapter covers:

- **The Logic of Hypothesis Testing** — the conceptual framework,
  errors, and p-values
- **One-Sample Tests** — testing a single mean against a known value
- **Two-Sample Tests** — comparing means between two independent groups
- **Paired Sample Test** — comparing means in matched or
  repeated-measures designs
- **One-Way ANOVA** — comparing means across three or more groups
- **Non-Parametric Alternatives** — robust tests when normality cannot
  be assumed
- **Effect Size and Statistical Power** — measuring practical
  significance and planning sample sizes

::: callout-note
## Learning Objectives

By the end of this chapter, you will be able to:

1.  Explain the logic of hypothesis testing including Type I and Type II
    errors.
2.  Select and apply the correct test for one-sample, two-sample, and
    paired designs.
3.  Conduct one-way ANOVA and interpret the F-statistic and post-hoc
    tests.
4.  Choose appropriate non-parametric alternatives when parametric
    assumptions are violated.
5.  Compute and interpret effect sizes and conduct power analyses for
    sample size planning.
6.  Report hypothesis test results in a format appropriate for academic
    research.
:::

------------------------------------------------------------------------

# The Logic of Hypothesis Testing

## Introduction

Before computing a single test statistic, it is essential to understand
*why* hypothesis testing works the way it does. The procedure is built
on a deceptively simple idea: assume the null hypothesis is true, then
ask how surprising the observed data would be under that assumption. If
the data are sufficiently unlikely under the null, we reject it. This
section establishes the conceptual framework that underlies every test
in this chapter and beyond.

## Theory

### The Null and Alternative Hypotheses

Every hypothesis test begins with two competing claims:

- **Null Hypothesis (**$H_0$): The "status quo" — a statement of no
  effect, no difference, or equality. It is the hypothesis we seek
  evidence *against*.
- **Alternative Hypothesis (**$H_1$ or $H_a$): The claim we seek
  evidence *for*. It represents the presence of an effect, difference,
  or relationship.

Hypotheses are always stated about **population parameters**, never
about sample statistics.

**Directionality of** $H_1$:

| Test Type    | $H_0$            | $H_1$            | Rejection Region |
|--------------|------------------|------------------|------------------|
| Two-tailed   | $\mu = \mu_0$    | $\mu \neq \mu_0$ | Both tails       |
| Left-tailed  | $\mu \geq \mu_0$ | $\mu < \mu_0$    | Left tail        |
| Right-tailed | $\mu \leq \mu_0$ | $\mu > \mu_0$    | Right tail       |

: Types of hypothesis tests

### The p-value

The **p-value** is the probability of observing a test statistic as
extreme as (or more extreme than) the one computed from the sample,
**assuming** $H_0$ is true.

$$p\text{-value} = P(\text{test statistic as extreme as observed} \mid H_0 \text{ is true})$$

A small p-value means the observed data are unlikely under $H_0$ —
providing evidence against it. A large p-value means the data are
consistent with $H_0$.

::: callout-warning
## What the p-value is NOT

- It is **not** $P(H_0 \text{ is true})$
- It is **not** the probability that the result occurred by chance
- It is **not** a measure of effect size or practical importance
- A statistically significant result is not necessarily practically
  important
:::

### The Significance Level $\alpha$

The **significance level** $\alpha$ is the threshold below which we
reject $H_0$. It represents the maximum acceptable probability of making
a Type I error. Common choices:

- $\alpha = 0.05$ — standard in most social and biological sciences
- $\alpha = 0.01$ — stricter; used when false positives are costly
- $\alpha = 0.001$ — very strict; used in genomics, particle physics

**Decision rule:** Reject $H_0$ if p-value $< \alpha$.

### Type I and Type II Errors

No decision rule is perfect. Two types of errors are possible:

|   | $H_0$ is True | $H_0$ is False |
|--------------|----------------------------|------------------------------|
| **Reject** $H_0$ | Type I Error ($\alpha$) | Correct ✓ (Power = $1-\beta$) |
| **Fail to Reject** $H_0$ | Correct ✓ | Type II Error ($\beta$) |

: Decision outcomes in hypothesis testing

- **Type I Error (False Positive):** Rejecting $H_0$ when it is actually
  true. Probability = $\alpha$, controlled by the researcher.
- **Type II Error (False Negative):** Failing to reject $H_0$ when $H_1$
  is true. Probability = $\beta$.
- **Power** = $1 - \beta$: the probability of correctly detecting a true
  effect.

There is an inherent tradeoff: decreasing $\alpha$ (stricter threshold)
reduces Type I errors but increases $\beta$ (more Type II errors). The
only way to reduce both simultaneously is to **increase the sample
size**.

### The General Procedure

Every hypothesis test follows the same five-step procedure:

1.  **State** $H_0$ and $H_1$ clearly, including directionality.
2.  **Choose** the significance level $\alpha$ and the appropriate test.
3.  **Compute** the test statistic from sample data.
4.  **Find** the p-value (or compare test statistic to critical value).
5.  **Decide** and **interpret** in context: reject or fail to reject
    $H_0$, and state the practical meaning.

### Confidence Intervals and Hypothesis Tests

Confidence intervals and hypothesis tests are two sides of the same
coin. A 95% confidence interval for $\mu$ contains all values of $\mu_0$
for which the two-tailed test at $\alpha = 0.05$ would **fail to
reject** $H_0: \mu = \mu_0$. When a CI excludes the null value, the
corresponding test rejects $H_0$. CIs are often more informative than
tests alone because they quantify the magnitude of the effect, not just
its significance.

## Example: The Logic of Hypothesis Testing

**Example 3.1.** A university claims that the average time for students
to complete a master's degree is 2 years (24 months). A researcher
suspects it takes longer and surveys 36 recent graduates, finding
$\bar{x} = 26.5$ months with $s = 7.2$ months.

**Step 1 — Hypotheses:**
$$H_0: \mu = 24 \qquad H_1: \mu > 24 \quad \text{(right-tailed)}$$

**Step 2 — Significance level:** $\alpha = 0.05$

**Step 3 — Test statistic** (t-test, $\sigma$ unknown):
$$t = \frac{\bar{x} - \mu_0}{s/\sqrt{n}} = \frac{26.5 - 24}{7.2/\sqrt{36}} = \frac{2.5}{1.2} = 2.083$$

**Step 4 — p-value** (right-tailed, df = 35):
$$p\text{-value} = P(T_{35} > 2.083) \approx 0.022$$

**Step 5 — Decision:** Since $p = 0.022 < \alpha = 0.05$, we **reject**
$H_0$.

**Interpretation:** There is statistically significant evidence at the
5% level that the true average completion time exceeds 24 months
($t_{35} = 2.08$, $p = 0.022$). The sample mean of 26.5 months suggests
students take approximately 2.5 months longer than the claimed average.

## R Example: Visualizing the p-value

```{r logic-viz}
# --- Visualize the p-value for Example 3.1 ---
df_val   <- 35
t_obs    <- 2.083
p_val    <- pt(t_obs, df = df_val, lower.tail = FALSE)

x_seq <- seq(-4, 4, length.out = 500)
t_df  <- data.frame(x = x_seq, y = dt(x_seq, df = df_val))

ggplot(t_df, aes(x, y)) +
  geom_line(color = "steelblue", linewidth = 1.2) +
  geom_area(data = subset(t_df, x >= t_obs),
            fill = "tomato", alpha = 0.6) +
  geom_vline(xintercept = t_obs, color = "tomato",
             linetype = "dashed", linewidth = 1) +
  annotate("text", x = 2.8, y = 0.25,
           label = paste0("p = ", round(p_val, 3)),
           color = "tomato", fontface = "bold", size = 4.5) +
  annotate("text", x = 2.083, y = -0.015,
           label = paste0("t = ", t_obs),
           color = "tomato", size = 3.8, hjust = 0.5) +
  labs(title    = "t-Distribution (df = 35): Right-Tailed Test",
       subtitle = "Shaded area = p-value = P(T > 2.083 | H₀ true)",
       x        = "t statistic", y = "Density") +
  theme_minimal(base_size = 13)
```

**Code explanation:**

- `dt()` computes the t-distribution density (PDF); `pt()` computes the
  CDF (cumulative probability).
- `pt(t_obs, df, lower.tail = FALSE)` gives the right-tail probability —
  the p-value for a right-tailed test.
- The shaded area visually represents the p-value: the probability of
  observing $t \geq 2.083$ if $H_0$ were true.

## Exercises

::: callout-tip
## Exercise 3.1

For each scenario below, state $H_0$ and $H_1$, identify the test
direction (two-tailed, left-, or right-tailed), and justify your choice:

(a) A pharmaceutical company claims its new drug reduces systolic blood
    pressure by at least 10 mmHg. A regulator wants to check if the
    reduction is less than claimed.
(b) A data scientist wants to know whether a new recommendation
    algorithm changes average session duration compared to the old
    algorithm (could be longer or shorter).
(c) An educator believes that students who attend extra tutorials score
    higher than those who do not.
:::

::: callout-tip
## Exercise 3.2

Explain in your own words why "failing to reject $H_0$" is not the same
as "proving $H_0$ is true." Use an analogy from everyday life to
illustrate the difference.
:::

------------------------------------------------------------------------

# One-Sample Tests

## Introduction

The **one-sample test** addresses the simplest inferential question:
does the population mean equal a specific hypothesized value $\mu_0$? It
is the natural starting point for learning hypothesis testing because it
involves a single group and a known reference value. In data science,
one-sample tests arise when benchmarking a system against a known
standard, verifying a manufacturer's claim, or checking whether a
model's accuracy exceeds chance.

## Theory

### One-Sample z-Test

Used when the **population standard deviation** $\sigma$ is known (rare
in practice) or the sample is very large ($n > 100$):

$$z = \frac{\bar{x} - \mu_0}{\sigma / \sqrt{n}} \sim N(0,1) \text{ under } H_0$$

Critical values: $z^* = \pm 1.96$ (two-tailed, $\alpha = 0.05$),
$z^* = 1.645$ (right-tailed, $\alpha = 0.05$).

### One-Sample t-Test

Used when $\sigma$ is **unknown** (estimated by $s$) — the typical case:

$$t = \frac{\bar{x} - \mu_0}{s / \sqrt{n}} \sim t(n-1) \text{ under } H_0$$

**Assumptions:**

1.  The sample is a random sample from the population.
2.  The variable is continuous (interval or ratio scale).
3.  The population is approximately normally distributed **or**
    $n \geq 30$ (CLT applies).

### Confidence Interval Approach

The $(1-\alpha) \times 100\%$ confidence interval for $\mu$ is:

$$\bar{x} \pm t^*_{n-1} \cdot \frac{s}{\sqrt{n}}$$

where $t^*_{n-1}$ is the critical value from the t-distribution with
$n-1$ degrees of freedom.

If $\mu_0$ falls **outside** this interval, reject $H_0$ at significance
level $\alpha$.

## Example: One-Sample t-Test

**Example 3.2.** A coffee machine is calibrated to dispense 250 ml per
cup. A quality control inspector measures 20 cups and finds
$\bar{x} = 246.8$ ml, $s = 7.3$ ml. Is there evidence the machine is
under-dispensing?

**Hypotheses:**
$$H_0: \mu = 250 \qquad H_1: \mu < 250 \quad \text{(left-tailed)}$$

**Test statistic:**
$$t = \frac{246.8 - 250}{7.3 / \sqrt{20}} = \frac{-3.2}{1.632} = -1.961$$

**p-value** (left-tailed, df = 19): $P(T_{19} < -1.961) = 0.032$

**Decision:** $p = 0.032 < \alpha = 0.05$ → **Reject** $H_0$.

**Interpretation:** There is significant evidence that the machine
dispenses less than 250 ml on average ($t_{19} = -1.96$, $p = 0.032$).
The one-sided 95% upper confidence bound for the true mean is 249.6 ml
— below the claimed 250 ml, consistent with rejecting $H_0$ in this
left-tailed test.

## R Example: One-Sample Tests

```{r one-sample}
# --- Simulate the coffee machine data ---
set.seed(101)
cups <- c(245.2, 236.0, 254.2, 248.6, 239.8, 257.3, 243.2,
          247.9, 255.9, 237.1, 250.7, 243.5, 238.0, 259.8,
          248.8, 242.1, 253.5, 236.2, 246.0, 252.3)

# One-sample t-test (left-tailed: H1: mu < 250)
t_result <- t.test(cups, mu = 250, alternative = "less")
print(t_result)
```

```{r one-sample-summary}
# --- Clean summary table ---
data.frame(
  Statistic   = c("Sample Mean", "Sample SD", "n",
                  "t statistic", "df", "p-value (left-tailed)",
                  "95% CI Lower", "95% CI Upper"),
  Value       = c(round(mean(cups), 2), round(sd(cups), 2), length(cups),
                  round(t_result$statistic, 3), t_result$parameter,
                  round(t_result$p.value, 4),
                  round(t_result$conf.int[1], 2),
                  round(t_result$conf.int[2], 2))
) |>
  kable(caption = "One-Sample t-Test Results: Coffee Machine",
        col.names = c("Statistic", "Value")) |>
  kable_styling(bootstrap_options = c("striped","hover"), full_width = FALSE)
```

```{r one-sample-plot}
# --- Visualize data with reference line ---
ggplot(data.frame(ml = cups), aes(x = ml)) +
  geom_histogram(bins = 8, fill = "steelblue",
                 color = "white", alpha = 0.8) +
  geom_vline(xintercept = mean(cups), color = "steelblue",
             linewidth = 1.2, linetype = "dashed") +
  geom_vline(xintercept = 250, color = "tomato",
             linewidth = 1.2, linetype = "solid") +
  annotate("text", x = mean(cups) - 0.5, y = 4.5,
           label = paste0("x̄ = ", round(mean(cups),1)),
           color = "steelblue", hjust = 1, size = 4) +
  annotate("text", x = 250.5, y = 4.5,
           label = "μ₀ = 250", color = "tomato",
           hjust = 0, size = 4) +
  labs(title    = "Coffee Machine Dispensing Volume",
       subtitle = "Sample mean (blue) vs. claimed mean (red)",
       x        = "Volume (ml)", y = "Count") +
  theme_minimal(base_size = 13)
```

**Code explanation:**

- `t.test(x, mu, alternative)` performs the one-sample t-test.
  `alternative` can be `"less"`, `"greater"`, or `"two.sided"`.
- The result object contains `$statistic` (t value), `$parameter` (df),
  `$p.value`, and `$conf.int`.
- The confidence interval returned by `t.test()` with
  `alternative = "less"` gives a one-sided upper bound — for a two-sided
  CI, use `alternative = "two.sided"`.

## Exercises

::: callout-tip
## Exercise 3.3

A manufacturer claims their batteries last 500 hours on average. You
test 25 batteries and find $\bar{x} = 487$ hours, $s = 42$ hours.

(a) State $H_0$ and $H_1$ for a two-tailed test.
(b) Compute the t-statistic and p-value by hand. Verify in R.
(c) Construct a 95% confidence interval for the true mean battery life.
(d) Based on both the test and the CI, what do you conclude?
:::

::: callout-tip
## Exercise 3.4

Using the `sleep` dataset built into R (extra sleep hours gained with
two drugs):

(a) Test whether the mean extra sleep for group 1 (`group == 1`) differs
    from zero (two-tailed, $\alpha = 0.05$).
(b) Test whether the mean extra sleep for group 2 (`group == 2`) is
    greater than zero (right-tailed).
(c) Interpret both results and compare conclusions.
:::

------------------------------------------------------------------------

# Two-Sample Tests

## Introduction

In many research and data science contexts, the question is not whether
a population mean equals a fixed value, but whether **two populations
have the same mean**. Does the treatment group differ from the control?
Do users of version A spend more time on the platform than users of
version B? The **two-sample t-test** is the standard tool for comparing
means from two independent groups, and it underlies A/B testing,
clinical trials, and comparative studies across all fields.

## Theory

### Independent Samples t-Test (Equal Variances)

When we assume the two populations have equal variances
($\sigma_1^2 = \sigma_2^2$), we pool the variance estimates:

$$s_p^2 = \frac{(n_1 - 1)s_1^2 + (n_2 - 1)s_2^2}{n_1 + n_2 - 2}$$

$$t = \frac{(\bar{x}_1 - \bar{x}_2) - \delta_0}{s_p\sqrt{\frac{1}{n_1} + \frac{1}{n_2}}} \sim t(n_1 + n_2 - 2) \text{ under } H_0$$

where $\delta_0$ is the hypothesized difference (usually 0).

### Welch's t-Test (Unequal Variances)

The equal-variance assumption is often untenable. **Welch's t-test**
does not assume equal variances and uses a modified degrees of freedom
(Satterthwaite approximation):

$$t = \frac{\bar{x}_1 - \bar{x}_2}{\sqrt{\frac{s_1^2}{n_1} + \frac{s_2^2}{n_2}}}$$

$$df \approx \frac{\left(\frac{s_1^2}{n_1} + \frac{s_2^2}{n_2}\right)^2}{\frac{(s_1^2/n_1)^2}{n_1-1} + \frac{(s_2^2/n_2)^2}{n_2-1}}$$

**Recommendation:** Welch's t-test is the default in R
(`var.equal = FALSE`) and is preferred in practice — it performs well
even when variances are equal, making it a safer choice.

### Testing for Equal Variances: Levene's Test

Before choosing between pooled and Welch's tests, some analysts check
whether variances are equal using **Levene's test** ($H_0$: equal
variances). However, many statisticians argue this pre-testing approach
inflates overall Type I error and recommend simply always using Welch's
test.

**Assumptions for both tests:**

1.  Independent random samples from each population.
2.  Continuous variable (interval or ratio scale).
3.  Approximately normal populations **or** $n_1, n_2 \geq 30$.

## Example: Two-Sample t-Test

**Example 3.3.** A researcher compares exam scores between two teaching
methods. Group A (traditional): $n_1 = 30$, $\bar{x}_1 = 74.2$,
$s_1 = 11.8$. Group B (flipped classroom): $n_2 = 28$,
$\bar{x}_2 = 81.6$, $s_2 = 9.4$.

**Hypotheses:**
$$H_0: \mu_A = \mu_B \qquad H_1: \mu_A \neq \mu_B \quad \text{(two-tailed)}$$

**Welch's t-statistic:**
$$t = \frac{74.2 - 81.6}{\sqrt{\frac{11.8^2}{30} + \frac{9.4^2}{28}}} = \frac{-7.4}{\sqrt{4.643 + 3.153}} = \frac{-7.4}{2.792} = -2.651$$

**df** (Satterthwaite) $\approx 55.6$, **p-value** $\approx 0.010$

**Decision:** $p = 0.010 < 0.05$ → **Reject** $H_0$.

**Interpretation:** Students in the flipped classroom scored
significantly higher than those in the traditional classroom
($t_{55.6} = -2.65$, $p = 0.010$). The mean difference of 7.4 points
(95% CI: 1.7 to 13.1) is both statistically significant and potentially
educationally meaningful.

## R Example: Two-Sample Tests

```{r two-sample}
# --- Simulate teaching method data ---
set.seed(202)
group_A <- rnorm(30, mean = 74.2, sd = 11.8)
group_B <- rnorm(28, mean = 81.6, sd = 9.4)

# Welch's t-test (default in R: var.equal = FALSE)
welch_result <- t.test(group_A, group_B,
                       alternative = "two.sided",
                       var.equal   = FALSE)

# Pooled t-test for comparison
pooled_result <- t.test(group_A, group_B,
                        alternative = "two.sided",
                        var.equal   = TRUE)

# Compare results
comparison <- data.frame(
  Test         = c("Welch's t-test", "Pooled t-test"),
  t_statistic  = round(c(welch_result$statistic,
                          pooled_result$statistic), 3),
  df           = round(c(welch_result$parameter,
                          pooled_result$parameter), 1),
  p_value      = round(c(welch_result$p.value,
                          pooled_result$p.value), 4),
  CI_lower     = round(c(welch_result$conf.int[1],
                          pooled_result$conf.int[1]), 2),
  CI_upper     = round(c(welch_result$conf.int[2],
                          pooled_result$conf.int[2]), 2)
)

kable(comparison,
      caption  = "Welch's vs. Pooled t-Test Results",
      col.names = c("Test", "t", "df", "p-value",
                    "95% CI Lower", "95% CI Upper")) |>
  kable_styling(bootstrap_options = c("striped","hover"), full_width = FALSE)
```

```{r two-sample-plot}
# --- Visualize group comparison ---
scores_df <- data.frame(
  score = c(group_A, group_B),
  group = rep(c("Group A\n(Traditional)", "Group B\n(Flipped)"),
              times = c(30, 28))
)

ggplot(scores_df, aes(x = group, y = score, fill = group)) +
  geom_violin(alpha = 0.5, trim = FALSE) +
  geom_boxplot(width = 0.15, fill = "white",
               outlier.shape = 21, outlier.size = 2) +
  stat_summary(fun = mean, geom = "point",
               shape = 23, size = 4, fill = "white") +
  scale_fill_manual(values = c("steelblue", "tomato")) +
  labs(title    = "Exam Scores by Teaching Method",
       subtitle = paste0("Welch's t-test: t = ",
                         round(welch_result$statistic, 2),
                         ", p = ", round(welch_result$p.value, 3)),
       x        = "Teaching Method",
       y        = "Exam Score") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none")
```

**Code explanation:**

- `t.test(x, y, alternative, var.equal)` handles both Welch's
  (`var.equal = FALSE`) and pooled (`var.equal = TRUE`) variants.
- `stat_summary(fun = mean, geom = "point", shape = 23)` adds a
  diamond-shaped mean marker inside the boxplot — a clean way to show
  both median (boxplot line) and mean simultaneously.
- The violin plot reveals distributional shape alongside the five-number
  summary, providing richer information than a boxplot alone.

## Exercises

::: callout-tip
## Exercise 3.5

Using the `mtcars` dataset, test whether automatic and manual
transmission cars have different fuel efficiency (`mpg`).

(a) Check normality for each group using Shapiro-Wilk and QQ plots.
(b) Run Welch's t-test. State hypotheses, report t, df, p-value, and 95%
    CI.
(c) Run the pooled t-test. Compare results to Welch's.
(d) Which test is more appropriate here? Justify using Levene's test
    (`var.test()` in R).
:::

::: callout-tip
## Exercise 3.6

A/B testing scenario: Two versions of a webpage are shown to random
users. Version A: $n = 200$, mean session = 4.2 min, $s = 1.8$ min.
Version B: $n = 200$, mean session = 4.7 min, $s = 2.1$ min.

(a) Conduct a two-tailed Welch's t-test at $\alpha = 0.05$. What do you
    conclude?
(b) Compute the 95% CI for the difference in means.
(c) Even if significant, should the company switch to Version B? What
    other factors matter?
:::

------------------------------------------------------------------------

# Paired Sample Test

## Introduction

The two-sample t-test assumes **independent** groups — the two samples
are drawn from completely separate populations. But many research
designs deliberately **match** or **pair** observations: measuring the
same subject before and after a treatment, comparing twins, or matching
patients by age and severity. In these designs, ignoring the pairing
wastes information and inflates variability. The **paired t-test**
exploits the within-pair relationship, making it more powerful than the
independent samples test when pairing is effective.

## Theory

### The Paired t-Test

Given $n$ pairs $(x_{1i}, x_{2i})$, compute the difference for each
pair:

$$d_i = x_{1i} - x_{2i}, \qquad i = 1, 2, \ldots, n$$

The paired t-test is simply a **one-sample t-test on the differences**
$d_i$, testing whether the mean difference $\mu_d$ equals zero:

$$H_0: \mu_d = 0 \qquad H_1: \mu_d \neq 0 \quad \text{(or one-sided)}$$

$$t = \frac{\bar{d}}{s_d / \sqrt{n}} \sim t(n-1) \text{ under } H_0$$

where $\bar{d}$ is the sample mean of differences and $s_d$ is the
standard deviation of differences.

**Key insight:** By working with differences, we eliminate
**between-subject variability** — a major source of noise in independent
samples designs. The paired test is more powerful when subjects are
heterogeneous but respond similarly to treatment.

**When to use paired vs. independent:**

| Use Paired t-Test               | Use Independent t-Test      |
|---------------------------------|-----------------------------|
| Same subject measured twice     | Two separate groups         |
| Matched pairs (twins, siblings) | Random assignment to groups |
| Before-after design             | Cross-sectional comparison  |

: Choosing between paired and independent t-tests

**Assumptions:**

1.  Pairs are randomly and independently selected.
2.  Differences $d_i$ are approximately normally distributed (not the
    raw data).
3.  The variable is continuous.

## Example: Paired t-Test

**Example 3.4.** Ten patients' systolic blood pressure (mmHg) is
measured before and after 8 weeks of a new medication:

| Patient | Before | After | Difference ($d_i$) |
|---------|--------|-------|--------------------|
| 1       | 145    | 132   | 13                 |
| 2       | 158    | 148   | 10                 |
| 3       | 162    | 151   | 11                 |
| 4       | 149    | 140   | 9                  |
| 5       | 155    | 143   | 12                 |
| 6       | 170    | 159   | 11                 |
| 7       | 152    | 147   | 5                  |
| 8       | 141    | 136   | 5                  |
| 9       | 163    | 150   | 13                 |
| 10      | 157    | 148   | 9                  |

$\bar{d} = 9.8$, $s_d = 2.90$, $n = 10$

**Hypotheses:** $H_0: \mu_d = 0$ vs. $H_1: \mu_d > 0$ (right-tailed —
testing for reduction)

**Test statistic:**
$$t = \frac{9.8}{2.90/\sqrt{10}} = \frac{9.8}{0.917} = 10.69$$

**p-value** (right-tailed, df = 9): $P(T_9 > 10.69) < 0.0001$

**Decision:** **Reject** $H_0$ with overwhelming evidence.

**Interpretation:** The medication produced a highly significant
reduction in systolic blood pressure ($t_9 = 10.69$, $p < 0.0001$). The
mean reduction of 9.8 mmHg (95% CI: 7.73 to 11.87 mmHg) is clinically
meaningful by conventional standards.

## R Example: Paired t-Test

```{r paired}
# --- Blood pressure data ---
before <- c(145, 158, 162, 149, 155, 170, 152, 141, 163, 157)
after  <- c(132, 148, 151, 140, 143, 159, 147, 136, 150, 148)
diff_d <- before - after

# Paired t-test
paired_result <- t.test(before, after,
                         paired      = TRUE,
                         alternative = "greater")

cat("Mean difference (before - after):", round(mean(diff_d), 2), "mmHg\n")
cat("SD of differences:               ", round(sd(diff_d), 2),  "mmHg\n")
print(paired_result)
```

```{r paired-plot}
# --- Visualize paired data ---
bp_df <- data.frame(
  patient = rep(1:10, 2),
  time    = rep(c("Before", "After"), each = 10),
  bp      = c(before, after)
)
bp_df$time <- factor(bp_df$time, levels = c("Before", "After"))

p1 <- ggplot(bp_df, aes(x = time, y = bp, group = patient)) +
  geom_line(color = "gray60", alpha = 0.7) +
  geom_point(aes(color = time), size = 3) +
  scale_color_manual(values = c("Before" = "tomato", "After" = "steelblue")) +
  labs(title    = "A. Individual Blood Pressure Changes",
       subtitle = "Each line represents one patient",
       x        = "Time Point", y = "Systolic BP (mmHg)") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none")

p2 <- ggplot(data.frame(d = diff_d), aes(x = d)) +
  geom_histogram(bins = 6, fill = "steelblue",
                 color = "white", alpha = 0.8) +
  geom_vline(xintercept = mean(diff_d), color = "tomato",
             linewidth = 1.2, linetype = "dashed") +
  geom_vline(xintercept = 0, color = "gray40",
             linewidth = 1, linetype = "solid") +
  annotate("text", x = mean(diff_d) + 0.3, y = 3.3,
           label = paste0("d̄ = ", round(mean(diff_d),1)),
           color = "tomato", size = 4) +
  labs(title    = "B. Distribution of Differences",
       subtitle = "Null value (0) vs. observed mean difference",
       x        = "Difference (Before − After)", y = "Count") +
  theme_minimal(base_size = 12)

p1 + p2
```

**Code explanation:**

- `t.test(x, y, paired = TRUE)` runs the paired t-test in R. Internally,
  it computes differences and runs a one-sample test.
- The spaghetti plot (Panel A) is the canonical visualization for paired
  data — each line traces one subject's trajectory, making
  individual-level change visible.
- Panel B shows the distribution of differences. The red dashed line
  (mean difference) being far from zero confirms the visual story of
  Panel A.

## Exercises

::: callout-tip
## Exercise 3.7

Use the built-in `sleep` dataset in R, which records extra hours of
sleep with two drugs for 10 patients.

(a) Treat this as paired data (same 10 patients in both groups). Compute
    differences (group 2 − group 1).
(b) Test $H_0: \mu_d = 0$ vs. $H_1: \mu_d \neq 0$ using
    `t.test(..., paired = TRUE)`.
(c) Produce a spaghetti plot and a histogram of differences.
(d) Would you reach a different conclusion using an independent samples
    t-test? Why does pairing matter here?
:::

------------------------------------------------------------------------

# One-Way ANOVA

## Introduction

The t-test compares means between **two** groups. But what if we have
three, four, or more groups? Conducting multiple pairwise t-tests is
statistically invalid — with $k$ groups, there are $\binom{k}{2}$
possible comparisons, and the probability of at least one false positive
increases rapidly (**familywise error rate inflation**). For $k = 5$
groups and $\alpha = 0.05$, running all 10 pairwise tests gives a
familywise error rate of $1 - 0.95^{10} \approx 40\%$. **One-way ANOVA**
solves this by testing all groups simultaneously in a single test.

## Theory

### The ANOVA Model

One-way ANOVA tests whether the means of $k$ independent groups are all
equal:

$$H_0: \mu_1 = \mu_2 = \cdots = \mu_k \qquad H_1: \text{at least one } \mu_i \neq \mu_j$$

The model decomposes each observation into components:

$$x_{ij} = \mu + \tau_i + \varepsilon_{ij}$$

where $\mu$ is the grand mean, $\tau_i$ is the effect of group $i$, and
$\varepsilon_{ij} \sim N(0, \sigma^2)$ is random error.

### Partitioning Variance

ANOVA works by partitioning the **total variability** in the data into
two sources:

$$SS_{\text{Total}} = SS_{\text{Between}} + SS_{\text{Within}}$$

| Source | Sum of Squares | df | Mean Square | F |
|-------------|-------------------|-------------|----------------|-------------|
| Between groups | $SS_B = \sum_i n_i(\bar{x}_i - \bar{x})^2$ | $k-1$ | $MS_B = SS_B/(k-1)$ | $MS_B/MS_W$ |
| Within groups | $SS_W = \sum_i\sum_j(x_{ij}-\bar{x}_i)^2$ | $N-k$ | $MS_W = SS_W/(N-k)$ |  |
| Total | $SS_T$ | $N-1$ |  |  |

: ANOVA table structure

The **F-statistic** is the ratio of between-group variability to
within-group variability:

$$F = \frac{MS_{\text{Between}}}{MS_{\text{Within}}} \sim F(k-1, N-k) \text{ under } H_0$$

If $H_0$ is true, $F \approx 1$ (between-group variation ≈ within-group
variation). A large $F$ suggests genuine group differences.

### Assumptions

1.  **Independence:** Observations are independent within and across
    groups.
2.  **Normality:** Observations within each group are approximately
    normally distributed.
3.  **Homogeneity of variance (homoscedasticity):**
    $\sigma_1^2 = \sigma_2^2 = \cdots = \sigma_k^2$. Test with Levene's
    test.

### Post-Hoc Tests

A significant ANOVA only tells us that **at least one** mean differs —
it does not tell us *which* pairs differ. **Post-hoc tests** conduct
pairwise comparisons while controlling the familywise error rate:

- **Tukey's HSD (Honest Significant Difference):** Most common; compares
  all pairs; controls familywise error at $\alpha$.
- **Bonferroni correction:** Divides $\alpha$ by the number of
  comparisons; conservative.
- **Scheffé's test:** Most flexible; controls error for all possible
  contrasts, not just pairwise.

## Example: One-Way ANOVA

**Example 3.5.** A nutritionist compares weight loss (kg) over 12 weeks
across three diet plans (A, B, C). Summary: $\bar{x}_A = 4.2$,
$\bar{x}_B = 6.8$, $\bar{x}_C = 5.5$, $n_i = 15$ each. The ANOVA F-test
gives $F(2, 42) = 8.43$, $p = 0.001$.

**Decision:** Reject $H_0$ — at least one diet plan produces different
mean weight loss.

**Post-hoc (Tukey HSD):** Diet B vs. A: $p = 0.001$ (significant); Diet
B vs. C: $p = 0.048$ (significant); Diet C vs. A: $p = 0.121$ (not
significant).

**Interpretation:** Diet B produces significantly greater weight loss
than both Diets A and C. Diets A and C do not differ significantly from
each other.

## R Example: One-Way ANOVA

```{r anova}
# --- Use the built-in PlantGrowth dataset ---
data(PlantGrowth)

# Descriptive statistics by group
PlantGrowth |>
  group_by(group) |>
  summarise(
    n      = n(),
    Mean   = round(mean(weight), 3),
    SD     = round(sd(weight), 3),
    Median = round(median(weight), 3)
  ) |>
  kable(caption = "Plant Weight by Treatment Group") |>
  kable_styling(bootstrap_options = c("striped","hover"), full_width = FALSE)
```

```{r anova-test}
# --- One-way ANOVA ---
anova_model  <- aov(weight ~ group, data = PlantGrowth)
anova_result <- summary(anova_model)
print(anova_result)
```

```{r anova-posthoc}
# --- Tukey HSD Post-Hoc Test ---
tukey_result <- TukeyHSD(anova_model)
print(tukey_result)

# Tidy post-hoc table
tukey_df <- as.data.frame(tukey_result$group)
tukey_df$Comparison <- rownames(tukey_df)
tukey_df <- tukey_df[, c(5, 1:4)]
tukey_df[, 2:5] <- round(tukey_df[, 2:5], 4)

kable(tukey_df,
      caption   = "Tukey HSD Post-Hoc Comparisons",
      row.names = FALSE,
      col.names = c("Comparison", "Difference",
                    "Lower CI", "Upper CI", "p adj")) |>
  kable_styling(bootstrap_options = c("striped","hover"), full_width = FALSE) |>
  column_spec(5, bold = TRUE,
              color = ifelse(tukey_df[, 5] < 0.05, "tomato", "gray40"))
```

```{r anova-plot}
# --- Visualization ---
p1 <- ggplot(PlantGrowth, aes(x = group, y = weight, fill = group)) +
  geom_boxplot(alpha = 0.7, outlier.shape = 21) +
  geom_jitter(width = 0.1, size = 2, alpha = 0.6) +
  scale_fill_brewer(palette = "Set2") +
  labs(title    = "A. Plant Weight by Treatment",
       x        = "Treatment Group",
       y        = "Dry Weight (g)") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none")

# Tukey HSD confidence intervals plot
p2 <- as.data.frame(tukey_result$group) |>
  rownames_to_column("comparison") |>
  ggplot(aes(x = comparison, y = diff,
             ymin = lwr, ymax = upr, color = comparison)) +
  geom_pointrange(size = 0.9) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray40") +
  scale_color_brewer(palette = "Set1") +
  coord_flip() +
  labs(title    = "B. Tukey HSD 95% CIs",
       subtitle = "CI crossing zero = not significant",
       x        = "Comparison",
       y        = "Mean Difference") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none")

p1 + p2
```

**Code explanation:**

- `aov(formula, data)` fits the ANOVA model; `summary()` produces the
  ANOVA table with F-statistic and p-value.
- `TukeyHSD(model)` performs all pairwise comparisons with familywise
  error control.
- The `pointrange` plot (Panel B) is the standard way to visualize Tukey
  HSD results — CIs crossing zero indicate non-significant differences.

## Exercises

::: callout-tip
## Exercise 3.8

Using the `iris` dataset, test whether mean `Sepal.Length` differs
across the three species.

(a) Check ANOVA assumptions: normality per group (Shapiro-Wilk) and
    equal variances (Levene's test via `car::leveneTest()`).
(b) Run one-way ANOVA. Report F, df, and p-value in APA format.
(c) Run Tukey HSD. Which species pairs differ significantly?
(d) Produce a boxplot and Tukey CI plot.
:::

::: callout-tip
## Exercise 3.9 (Challenge)

The `npk` dataset in R contains crop yield data from a factorial
experiment.

(a) Use one-way ANOVA to test whether nitrogen treatment (`N`) affects
    yield.
(b) What happens to your conclusions if you violate the homoscedasticity
    assumption? Rerun the analysis using Welch's ANOVA
    (`oneway.test(var.equal = FALSE)`) and compare.
:::

------------------------------------------------------------------------

# Non-Parametric Alternatives

## Introduction

Parametric tests (t-tests, ANOVA) assume normally distributed
populations and, in some cases, equal variances. When these assumptions
are seriously violated — particularly with small samples, heavily skewed
data, or ordinal outcomes — **non-parametric tests** provide valid
alternatives. These tests make no assumption about the distributional
form of the data, instead working with **ranks** rather than raw values.
The cost is a small loss of statistical power when parametric
assumptions are actually met.

## Theory

### Wilcoxon Signed-Rank Test (One-Sample / Paired)

The **non-parametric alternative to the one-sample and paired t-tests**.
It tests whether the median of differences equals zero, without assuming
normality.

**Procedure:**

1.  Compute differences $d_i = x_{1i} - x_{2i}$ (or $d_i = x_i - \mu_0$
    for one-sample).
2.  Rank the **absolute values** $|d_i|$, ignoring zeros.
3.  Assign signs to ranks based on the sign of $d_i$.
4.  Compute $W^+$ (sum of positive ranks) and $W^-$ (sum of negative
    ranks).
5.  Test statistic: $W = \min(W^+, W^-)$.

For large samples, $W$ is approximately normal and a z-approximation is
used.

### Mann-Whitney U Test (Two Independent Samples)

The **non-parametric alternative to the independent samples t-test**.
Tests whether one population tends to have larger values than another
(stochastic dominance), without assuming normality.

**Procedure:**

1.  Pool and rank all observations from both groups.
2.  Compute $U_1$ (sum of ranks from group 1 minus the minimum possible)
    and $U_2$.
3.  Test statistic: $U = \min(U_1, U_2)$.

In R, both the Wilcoxon signed-rank and Mann-Whitney U tests are
implemented in `wilcox.test()`.

### Kruskal-Wallis Test (Three or More Groups)

The **non-parametric alternative to one-way ANOVA**. Tests whether the
distributions of $k$ groups are identical (with the alternative that at
least one group tends to have different values):

$$H = \frac{12}{N(N+1)} \sum_{i=1}^{k} \frac{R_i^2}{n_i} - 3(N+1) \sim \chi^2(k-1) \text{ under } H_0$$

where $R_i$ is the sum of ranks in group $i$ and $N$ is the total sample
size.

**Post-hoc for Kruskal-Wallis:** Dunn's test with Bonferroni or
Benjamini-Hochberg correction.

### When to Choose Non-Parametric Tests

| Situation | Non-Parametric Test | Parametric Equivalent |
|-----------------|--------------------------|-----------------------------|
| One sample or paired, non-normal | Wilcoxon signed-rank | Paired t-test |
| Two independent groups, non-normal | Mann-Whitney U | Independent t-test |
| Three or more groups, non-normal | Kruskal-Wallis | One-way ANOVA |
| Small $n$ (\< 15 per group) with skew | Any of the above | Requires normal assumption |
| Ordinal outcome variable | Any of the above | Inappropriate |

: Non-parametric test selection guide

## Example: Mann-Whitney U Test

**Example 3.6.** A researcher compares reaction times (ms) between
younger ($n_1 = 12$) and older ($n_2 = 10$) adults. The older group
shows strong right skew (skewness = 2.1), violating normality. The
Mann-Whitney U test gives $U = 18$, $p = 0.003$.

**Interpretation:** Older adults have significantly longer reaction
times than younger adults ($W = 18$, $p = 0.003$). The non-parametric
test was appropriate given the violation of normality in the older
group.

## R Example: Non-Parametric Tests

```{r nonparam}
# --- Use airquality: compare Ozone in May vs. August ---
data(airquality)
ozone_may <- na.omit(airquality$Ozone[airquality$Month == 5])
ozone_aug <- na.omit(airquality$Ozone[airquality$Month == 8])

# Check normality
cat("Shapiro-Wilk — May Ozone:    p =",
    round(shapiro.test(ozone_may)$p.value, 4), "\n")
cat("Shapiro-Wilk — August Ozone: p =",
    round(shapiro.test(ozone_aug)$p.value, 4), "\n\n")

# Mann-Whitney U test (non-parametric two-sample)
mw_result <- wilcox.test(ozone_may, ozone_aug,
                          alternative = "two.sided",
                          exact       = FALSE)
print(mw_result)

# Compare: Welch's t-test on the same data
t_result_ozone <- t.test(ozone_may, ozone_aug)
cat("\nWelch's t-test p-value:     ", round(t_result_ozone$p.value, 4), "\n")
cat("Mann-Whitney U p-value:     ", round(mw_result$p.value, 4), "\n")
```

```{r nonparam-kw}
# --- Kruskal-Wallis: Ozone across all months ---
airquality_kw <- na.omit(airquality[, c("Ozone","Month")])

kw_result <- kruskal.test(Ozone ~ factor(Month), data = airquality_kw)
print(kw_result)
```

```{r nonparam-plot}
# --- Visualization ---
month_labels <- c("5" = "May", "6" = "Jun",
                  "7" = "Jul", "8" = "Aug", "9" = "Sep")

airquality_kw |>
  mutate(Month = factor(Month, labels = month_labels)) |>
  ggplot(aes(x = Month, y = Ozone, fill = Month)) +
  geom_boxplot(alpha = 0.7, outlier.shape = 21) +
  geom_jitter(width = 0.1, size = 1.8, alpha = 0.5) +
  scale_fill_brewer(palette = "Set2") +
  labs(title    = "Ozone Concentration by Month",
       subtitle = paste0("Kruskal-Wallis: χ² = ",
                         round(kw_result$statistic, 2),
                         ", df = ", kw_result$parameter,
                         ", p = ", round(kw_result$p.value, 4)),
       x        = "Month", y = "Ozone (ppb)") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none")
```

**Code explanation:**

- `wilcox.test(x, y, alternative, exact)` runs both the Mann-Whitney U
  (two independent samples) and Wilcoxon signed-rank (one-sample or
  paired with `paired = TRUE`) test.
- `exact = FALSE` uses a normal approximation — appropriate when there
  are ties in the data.
- `kruskal.test(formula, data)` runs the Kruskal-Wallis test. For
  post-hoc, use `dunn_test()` from `rstatix`.

## Exercises

::: callout-tip
## Exercise 3.10

Using the `sleep` dataset (extra sleep hours for two drugs, 10 patients
each):

(a) Test normality of differences. Is the paired t-test appropriate?
(b) Run the Wilcoxon signed-rank test
    (`wilcox.test(..., paired = TRUE)`).
(c) Compare results with the paired t-test from Exercise 3.7. Do
    conclusions agree?
(d) Under what circumstances would the two tests disagree?
:::

::: callout-tip
## Exercise 3.11

Using the `iris` dataset, test whether `Petal.Length` differs across
species using the Kruskal-Wallis test.

(a) State why you might prefer Kruskal-Wallis over ANOVA for this
    variable.
(b) Run `kruskal.test()` and interpret the result.
(c) Perform Dunn's post-hoc test using `rstatix::dunn_test()` with
    Bonferroni correction.
(d) Do the conclusions match those from the parametric ANOVA in Exercise
    3.8?
:::

------------------------------------------------------------------------

# Effect Size and Statistical Power

## Introduction

A statistically significant result answers only one question: *Is the
effect real (not due to chance)?* It says nothing about whether the
effect is **practically important**. With a large enough sample, even a
trivially small difference becomes statistically significant. **Effect
size** measures the magnitude of an effect independently of sample size,
answering: *How big is this?* **Statistical power** determines whether
our study is large enough to detect an effect if one truly exists.
Together, effect size and power analysis are essential tools for
designing rigorous research and interpreting published results.

## Theory

### Effect Size Measures

**Cohen's d** (for mean comparisons):
$$d = \frac{\bar{x}_1 - \bar{x}_2}{s_{\text{pooled}}}$$

| Cohen's d | Interpretation |
|-----------|----------------|
| 0.2       | Small effect   |
| 0.5       | Medium effect  |
| 0.8       | Large effect   |

: Cohen's d benchmarks

$\eta^2$ (Eta-squared) for ANOVA:
$$\eta^2 = \frac{SS_{\text{Between}}}{SS_{\text{Total}}}$$

Represents the proportion of total variance explained by group
membership. Values of 0.01, 0.06, and 0.14 correspond to small, medium,
and large effects respectively.

$r$ (Correlation coefficient) for non-parametric tests:
$$r = \frac{Z}{\sqrt{N}}$$

where $Z$ is the standardized test statistic. Values of 0.1, 0.3, and
0.5 are small, medium, and large.

### Statistical Power

**Power** = $P(\text{Reject } H_0 \mid H_1 \text{ is true}) = 1 - \beta$

Power depends on four interrelated factors:

| Factor                        | Effect on Power |
|-------------------------------|-----------------|
| Sample size $n$ ↑             | Power ↑         |
| Effect size $d$ ↑             | Power ↑         |
| Significance level $\alpha$ ↑ | Power ↑         |
| Variability $\sigma$ ↑        | Power ↓         |

: Factors affecting statistical power

**Conventional target:** Power $\geq 0.80$ (80%) — meaning we have at
least an 80% chance of detecting a true effect.

### Power Analysis for Sample Size Planning

Power analysis is performed **before data collection** to determine the
required sample size. Given:

- The desired power ($1 - \beta$, typically 0.80 or 0.90)
- The significance level ($\alpha$, typically 0.05)
- The minimum effect size of practical interest ($d$ or $f$)

We solve for $n$. This is a critical step in research design —
under-powered studies waste resources and fail to detect real effects;
over-powered studies waste money and may detect trivially small effects.

## Example: Effect Size and Power

**Example 3.7.** Returning to the teaching method comparison (Example
3.3):

$$d = \frac{|81.6 - 74.2|}{\sqrt{(11.8^2 + 9.4^2)/2}} = \frac{7.4}{\sqrt{113.8}} = \frac{7.4}{10.67} \approx 0.69$$

A Cohen's d of 0.69 indicates a **medium-to-large** effect. The
difference of 7.4 points is both statistically significant and
educationally meaningful. A researcher planning a replication study who
wants 80% power to detect $d = 0.69$ at $\alpha = 0.05$ (two-tailed)
would need approximately $n = 32$ per group.

## R Example: Effect Size and Power Analysis

```{r effect-size}
# --- Cohen's d for teaching methods ---
library(effectsize)

set.seed(202)
group_A <- rnorm(30, mean = 74.2, sd = 11.8)
group_B <- rnorm(28, mean = 81.6, sd = 9.4)

# Cohen's d
d_result <- effectsize::cohens_d(group_B, group_A)
print(d_result)
interpret_cohens_d(d_result$Cohens_d)
```

```{r power-analysis}
# --- Power analysis using pwr package ---
library(pwr)

# 1. What power do we have with n=30 per group and d=0.69?
power_current <- pwr.t.test(n    = 30,
                             d    = 0.69,
                             sig.level = 0.05,
                             type = "two.sample",
                             alternative = "two.sided")
cat("Power with n=30, d=0.69:", round(power_current$power, 3), "\n\n")

# 2. What n do we need for 80% power with d=0.69?
n_needed <- pwr.t.test(power = 0.80,
                        d     = 0.69,
                        sig.level   = 0.05,
                        type        = "two.sample",
                        alternative = "two.sided")
cat("n needed for 80% power, d=0.69:", ceiling(n_needed$n), "per group\n\n")

# 3. What n do we need for 90% power?
n_needed_90 <- pwr.t.test(power = 0.90,
                           d     = 0.69,
                           sig.level   = 0.05,
                           type        = "two.sample",
                           alternative = "two.sided")
cat("n needed for 90% power, d=0.69:", ceiling(n_needed_90$n), "per group\n")
```

```{r power-curve}
# --- Power curve: power vs. sample size for different effect sizes ---
n_seq     <- seq(10, 100, by = 2)
d_values  <- c(0.2, 0.5, 0.8)
d_labels  <- c("Small (d=0.2)", "Medium (d=0.5)", "Large (d=0.8)")

power_df <- map_dfr(seq_along(d_values), function(i) {
  d   <- d_values[i]
  lbl <- d_labels[i]
  pwr_vals <- sapply(n_seq, function(n) {
    pwr.t.test(n = n, d = d, sig.level = 0.05,
               type = "two.sample",
               alternative = "two.sided")$power
  })
  data.frame(n = n_seq, power = pwr_vals, effect = lbl)
})

ggplot(power_df, aes(x = n, y = power, color = effect)) +
  geom_line(linewidth = 1.2) +
  geom_hline(yintercept = 0.80, linetype = "dashed",
             color = "gray30", linewidth = 0.8) +
  geom_hline(yintercept = 0.90, linetype = "dotted",
             color = "gray30", linewidth = 0.8) +
  annotate("text", x = 95, y = 0.82,
           label = "80% power", color = "gray30", size = 3.5) +
  annotate("text", x = 95, y = 0.92,
           label = "90% power", color = "gray30", size = 3.5) +
  scale_color_manual(values = c("steelblue","seagreen","tomato")) +
  scale_y_continuous(labels = scales::percent) +
  labs(title    = "Power Curves: Two-Sample t-Test (α = 0.05)",
       subtitle = "Each curve shows a different effect size",
       x        = "Sample Size per Group (n)",
       y        = "Statistical Power",
       color    = "Effect Size") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "top")
```

**Code explanation:**

- `cohens_d()` from `effectsize` computes Cohen's d with confidence
  intervals; `interpret_cohens_d()` gives the verbal label
  (small/medium/large).
- `pwr.t.test()` from `pwr` performs power analysis for t-tests. Set any
  three of (n, d, sig.level, power) and it solves for the fourth.
- The power curve plot is an essential tool for research design — it
  visually shows the sample size required to achieve desired power for
  effects of different magnitudes.

## Exercises

::: callout-tip
## Exercise 3.12

For the `mtcars` analysis from Exercise 3.5 (mpg by transmission type):

(a) Compute Cohen's d for the mean difference between automatic and
    manual transmission cars.
(b) Interpret the effect size using Cohen's benchmarks.
(c) How many cars per group would be needed to achieve 80% power to
    detect this effect at $\alpha = 0.05$?
(d) Plot a power curve showing power vs. sample size for small, medium,
    and the observed effect size.
:::

::: callout-tip
## Exercise 3.13 (Challenge)

A researcher is planning a three-arm clinical trial comparing a new
drug, an existing drug, and placebo. They expect a medium effect size
($f = 0.25$ for ANOVA) and want 80% power at $\alpha = 0.05$.

(a) Use `pwr.anova.test()` to determine the required sample size per
    group.
(b) How does the required n change if power is increased to 90%?
(c) Compute $\eta^2$ from the `PlantGrowth` ANOVA result and interpret
    its magnitude.
:::

------------------------------------------------------------------------

# Chapter Lab Activity: Exploring Hypothesis Testing with the `ToothGrowth` Dataset

## Objectives

In this lab you will apply the full hypothesis testing workflow to the
`ToothGrowth` dataset — a classic R dataset recording the length of
odontoblasts (cells responsible for tooth growth) in 60 guinea pigs,
each receiving one of three doses of Vitamin C (0.5, 1, and 2 mg/day)
via one of two delivery methods (orange juice or ascorbic acid). You
will move from checking assumptions to selecting tests, reporting
results, and computing effect sizes.

## The Dataset

```{r lab-intro}
data(ToothGrowth)

kable(head(ToothGrowth, 12),
      caption = "First 12 Rows of the ToothGrowth Dataset") |>
  kable_styling(bootstrap_options = c("striped","hover"), font_size = 11)

cat("Dimensions:", nrow(ToothGrowth), "rows x",
    ncol(ToothGrowth), "columns\n")
cat("Supplement types:", levels(factor(ToothGrowth$supp)), "\n")
cat("Dose levels:", unique(ToothGrowth$dose), "mg/day\n")
```

**Variable descriptions:**

| Variable | Description | Scale |
|------------------------|-------------------------------|------------------|
| `len` | Odontoblast length (microns) | Ratio, continuous |
| `supp` | Supplement type: OJ (orange juice) or VC (ascorbic acid) | Nominal |
| `dose` | Vitamin C dose (0.5, 1.0, 2.0 mg/day) | Ordinal (treated as ratio here) |

## Lab Task 1: Descriptive Overview and Assumption Checks

```{r lab-task1}
# Descriptive statistics by supplement and dose
ToothGrowth |>
  mutate(dose = factor(dose)) |>
  group_by(supp, dose) |>
  summarise(
    n      = n(),
    Mean   = round(mean(len), 2),
    SD     = round(sd(len), 2),
    Median = round(median(len), 2),
    .groups = "drop"
  ) |>
  kable(caption = "Tooth Length by Supplement and Dose") |>
  kable_styling(bootstrap_options = c("striped","hover"), full_width = FALSE)
```

```{r lab-task1b}
# Normality check per group
ToothGrowth |>
  group_by(supp) |>
  summarise(
    SW_W       = round(shapiro.test(len)$statistic, 4),
    SW_pvalue  = round(shapiro.test(len)$p.value, 4),
    Skewness   = round(moments::skewness(len), 3),
    .groups    = "drop"
  ) |>
  kable(caption = "Normality Assessment by Supplement Type") |>
  kable_styling(bootstrap_options = c("striped","hover"), full_width = FALSE)
```

## Lab Task 2: Two-Sample Test — OJ vs. VC

```{r lab-task2}
# Split by supplement
oj <- ToothGrowth$len[ToothGrowth$supp == "OJ"]
vc <- ToothGrowth$len[ToothGrowth$supp == "VC"]

# Welch's t-test
t_supp <- t.test(oj, vc, alternative = "two.sided")

# Mann-Whitney U (non-parametric comparison)
mw_supp <- wilcox.test(oj, vc, alternative = "two.sided", exact = FALSE)

# Cohen's d
d_supp <- effectsize::cohens_d(oj, vc)

results_supp <- data.frame(
  Test       = c("Welch's t-test", "Mann-Whitney U"),
  Statistic  = round(c(t_supp$statistic, mw_supp$statistic), 3),
  p_value    = round(c(t_supp$p.value, mw_supp$p.value), 4),
  Decision   = ifelse(c(t_supp$p.value, mw_supp$p.value) < 0.05,
                      "Reject H₀", "Fail to Reject H₀")
)

kable(results_supp,
      caption   = "OJ vs. VC: Test Results at α = 0.05",
      col.names = c("Test", "Statistic", "p-value", "Decision")) |>
  kable_styling(bootstrap_options = c("striped","hover"), full_width = FALSE) |>
  column_spec(4, bold = TRUE,
              color = ifelse(results_supp$Decision == "Reject H₀",
                             "tomato", "steelblue"))

cat("\nCohen's d:", round(d_supp$Cohens_d, 3),
    "(", interpret_cohens_d(d_supp$Cohens_d), ")\n")

```

## Lab Task 3: One-Way ANOVA — Effect of Dose

```{r lab-task3}
# One-way ANOVA: does dose affect tooth length?
ToothGrowth$dose_f <- factor(ToothGrowth$dose)
anova_dose   <- aov(len ~ dose_f, data = ToothGrowth)
tukey_dose   <- TukeyHSD(anova_dose)

cat("=== ANOVA: Tooth Length by Dose ===\n")
print(summary(anova_dose))

cat("\n=== Tukey HSD Post-Hoc ===\n")
print(tukey_dose)

# Eta-squared
ss_between <- summary(anova_dose)[[1]][1,"Sum Sq"]
ss_total   <- sum(summary(anova_dose)[[1]][,"Sum Sq"])
eta_sq     <- ss_between / ss_total
cat("\nEta-squared (η²):", round(eta_sq, 4),
    "— proportion of variance explained by dose\n")
```

## Lab Task 4: Comprehensive Visualization

```{r lab-task4, fig.height=8}
# Panel A: Overall distribution by supplement
p1 <- ggplot(ToothGrowth, aes(x = supp, y = len, fill = supp)) +
  geom_violin(alpha = 0.5, trim = FALSE) +
  geom_boxplot(width = 0.12, fill = "white",
               outlier.shape = 21) +
  scale_fill_manual(values = c("OJ" = "tomato", "VC" = "steelblue")) +
  labs(title = "A. Length by Supplement",
       x = "Supplement", y = "Tooth Length (µm)") +
  theme_minimal(base_size = 11) +
  theme(legend.position = "none")

# Panel B: Length by dose
p2 <- ggplot(ToothGrowth,
             aes(x = factor(dose), y = len, fill = factor(dose))) +
  geom_boxplot(alpha = 0.7, outlier.shape = 21) +
  geom_jitter(width = 0.1, size = 1.8, alpha = 0.5) +
  scale_fill_brewer(palette = "Blues") +
  labs(title = "B. Length by Dose",
       x = "Dose (mg/day)", y = "Tooth Length (µm)") +
  theme_minimal(base_size = 11) +
  theme(legend.position = "none")

# Panel C: Interaction — supplement × dose
p3 <- ToothGrowth |>
  group_by(supp, dose) |>
  summarise(mean_len = mean(len), se = sd(len)/sqrt(n()), .groups="drop") |>
  ggplot(aes(x = factor(dose), y = mean_len,
             group = supp, color = supp)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 3.5) +
  geom_errorbar(aes(ymin = mean_len - se, ymax = mean_len + se),
                width = 0.1) +
  scale_color_manual(values = c("OJ" = "tomato", "VC" = "steelblue")) +
  labs(title = "C. Interaction: Supplement × Dose",
       x = "Dose (mg/day)", y = "Mean Tooth Length (µm)",
       color = "Supplement") +
  theme_minimal(base_size = 11)

# Panel D: Tukey HSD intervals for dose
p4 <- as.data.frame(tukey_dose$dose_f) |>
  rownames_to_column("comparison") |>
  ggplot(aes(x = comparison, y = diff,
             ymin = lwr, ymax = upr,
             color = diff > 0)) +
  geom_pointrange(size = 0.9) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  scale_color_manual(values = c("FALSE" = "steelblue",
                                "TRUE"  = "tomato")) +
  coord_flip() +
  labs(title = "D. Tukey HSD: Dose Comparisons",
       x = "Comparison", y = "Mean Difference") +
  theme_minimal(base_size = 11) +
  theme(legend.position = "none")

(p1 + p2) / (p3 + p4)
```

## Lab Discussion Questions

Answer the following in writing:

1.  **Test Selection:** In Lab Task 2, both Welch's t-test and
    Mann-Whitney U give similar conclusions. Under what conditions might
    they disagree? Which would you report in a paper, and why?

2.  **Practical Significance:** The ANOVA in Task 3 is highly
    significant. Compute and interpret $\eta^2$. Does the dose variable
    explain a large, medium, or small proportion of variance in tooth
    length?

3.  **Post-Hoc Interpretation:** Based on Tukey HSD, which dose
    transition produces the greatest gain in tooth length: 0.5 → 1.0
    mg/day or 1.0 → 2.0 mg/day? What does this suggest about
    dose-response?

4.  **Interaction Pattern:** Panel C shows that at low doses, OJ
    outperforms VC, but at 2.0 mg/day the two converge. What does this
    interaction pattern imply for designing a Vitamin C supplementation
    study?

5.  **Power and Replication:** With only 10 animals per supplement-dose
    combination, compute the power of the two-sample t-test to detect
    the observed effect size between OJ and VC. Is this study adequately
    powered?

------------------------------------------------------------------------

# Chapter Summary

## Summary

This chapter established hypothesis testing as a principled framework
for data-driven decision making:

- **The logic of hypothesis testing** — stating $H_0$ and $H_1$,
  computing p-values, controlling Type I and II errors — is the same
  across all tests; only the test statistic changes.
- **One-sample t-tests** compare a single population mean to a
  hypothesized value.
- **Two-sample tests** (Welch's t-test preferred) compare means between
  two independent groups.
- **Paired t-tests** exploit within-subject pairing to eliminate
  between-subject noise and increase power.
- **One-way ANOVA** extends t-tests to three or more groups, avoiding
  familywise error inflation; post-hoc tests (Tukey HSD) identify which
  pairs differ.
- **Non-parametric tests** (Wilcoxon, Mann-Whitney, Kruskal-Wallis)
  provide valid alternatives when normality or equal-variance
  assumptions are violated.
- **Effect size** (Cohen's d, $\eta^2$) measures practical importance
  independently of sample size; **power analysis** ensures studies are
  designed to detect meaningful effects.

::: callout-important
## Key Formulas to Know

**One-sample t-statistic:**
$$t = \frac{\bar{x} - \mu_0}{s/\sqrt{n}} \sim t(n-1)$$

**Two-sample t-statistic (Welch's):**
$$t = \frac{\bar{x}_1 - \bar{x}_2}{\sqrt{s_1^2/n_1 + s_2^2/n_2}}$$

**Paired t-statistic:** $$t = \frac{\bar{d}}{s_d/\sqrt{n}} \sim t(n-1)$$

**F-statistic (ANOVA):**
$$F = \frac{MS_{\text{Between}}}{MS_{\text{Within}}} \sim F(k-1, N-k)$$

**Cohen's d:** $$d = \frac{\bar{x}_1 - \bar{x}_2}{s_{\text{pooled}}}$$

**Eta-squared:**
$$\eta^2 = \frac{SS_{\text{Between}}}{SS_{\text{Total}}}$$
:::

## References

Anderson, E. (1935). The irises of the Gaspé Peninsula. *Bulletin of the American Iris Society*, *59*, 2–5.

Cohen, J. (1988). *Statistical power analysis for the behavioral sciences* (2nd ed.). Lawrence Erlbaum Associates.

Fisher, R. A. (1925). *Statistical methods for research workers*. Oliver and Boyd.

Fox, J., & Weisberg, S. (2019). *An R companion to applied regression* (3rd ed.). SAGE.

Henderson, H. V., & Velleman, P. F. (1981). Building multiple regression models interactively. *Biometrics*, *37*(2), 391–411.

Kassambara, A. (2023). *rstatix: Pipe-friendly framework for basic statistical tests* (R package version 0.7.2). https://CRAN.R-project.org/package=rstatix

Lüdecke, D., Ben-Shachar, M. S., Patil, I., & Makowski, D. (2020). Extracting, computing and exploring the parameters of statistical models using R. *Journal of Open Source Software*, *5*(53), 2445.

Mann, H. B., & Whitney, D. R. (1947). On a test of whether one of two random variables is stochastically larger than the other. *Annals of Mathematical Statistics*, *18*(1), 50–60.

Pedersen, T. L. (2022). *patchwork: The composer of plots* (R package version 1.1.2). https://CRAN.R-project.org/package=patchwork

R Core Team. (2024). *R: A language and environment for statistical computing*. R Foundation for Statistical Computing. https://www.R-project.org/

Shapiro, S. S., & Wilk, M. B. (1965). An analysis of variance test for normality. *Biometrika*, *52*(3/4), 591–611.

Tukey, J. W. (1977). *Exploratory data analysis*. Addison-Wesley.

Venables, W. N., & Ripley, B. D. (2002). *Modern applied statistics with S* (4th ed.). Springer.

Wickham, H. (2016). *ggplot2: Elegant graphics for data analysis*. Springer.

Wickham, H., Averick, M., Bryan, J., Chang, W., McGowan, L. D., François, R., Grolemund, G., Hayes, A., Henry, L., Hester, J., Kuhn, M., Pedersen, T. L., Miller, E., Bache, S. M., Müller, K., Ooms, J., Robinson, D., Seidel, D. P., Spinu, V., … Yutani, H. (2019). Welcome to the tidyverse. *Journal of Open Source Software*, *4*(43), 1686.

Wilcoxon, F. (1945). Individual comparisons by ranking methods. *Biometrics Bulletin*, *1*(6), 80–83.

Xie, Y. (2015). *Dynamic documents with R and knitr* (2nd ed.). CRC Press.

------------------------------------------------------------------------

*End of Chapter 3. Proceed to Chapter 4: Test of Independence of
Variables.*
