# Chapter 2: Data Distribution and Probability

```{r setup}
# Load all libraries needed for this chapter
library(tidyverse)
library(knitr)
library(kableExtra)
library(patchwork)
library(nortest)      # normality tests (ad.test, lillie.test)
```

---

# Chapter Overview

## Introduction

Understanding how data is distributed — and the probability rules that govern those distributions — is the theoretical backbone of all statistical inference. Before we can test hypotheses, build models, or make predictions, we must know what kind of data we have, what probability laws apply to it, and how random variation behaves.

This chapter covers:

- **Types of Data and Measurement Scales** — what kind of variable are we dealing with?
- **Probability Fundamentals** — the rules of reasoning under uncertainty
- **Conditional Probability and Bayes' Theorem** — updating beliefs with evidence
- **Discrete Probability Distributions** — Binomial and Poisson
- **Continuous Probability Distributions** — Normal, Uniform, and Exponential
- **Sampling Distributions and the Central Limit Theorem** — why the normal distribution is everywhere
- **Assessing Normality** — how to test whether data is normally distributed

::: {.callout-note}
## Learning Objectives
By the end of this chapter, you will be able to:

1. Classify variables by measurement scale and identify the appropriate analysis for each.
2. Apply the fundamental rules of probability to solve real problems.
3. Use Bayes' Theorem to update probabilities based on new evidence.
4. Compute probabilities from discrete and continuous distributions by hand and in R.
5. Explain the Central Limit Theorem and its implications for statistical inference.
6. Assess normality using graphical and formal test methods in R.
:::

---

# Types of Data and Measurement Scales

## Introduction

Not all data is the same. A zip code, a temperature reading, a Likert-scale rating, and a weight measurement may all look like numbers — but they carry fundamentally different amounts of information and support different mathematical operations. Misidentifying the scale of measurement is one of the most common errors in data analysis, leading to inappropriate statistical methods and misleading conclusions.

The **Stevens classification system** (1946) organizes variables into four hierarchical levels of measurement, each building on the properties of the one before it.

## Theory

### Nominal Scale

Nominal variables classify observations into **mutually exclusive, unordered categories**. Numbers assigned to nominal categories are labels only — arithmetic operations are meaningless.

- **Properties:** Identity only (categories are distinct)
- **Examples:** Blood type (A, B, AB, O), country of origin, species, gender
- **Permissible statistics:** Frequency counts, mode, chi-square tests
- **Not permissible:** Mean, median, standard deviation

### Ordinal Scale

Ordinal variables have categories with a **meaningful order or ranking**, but the intervals between ranks are not necessarily equal.

- **Properties:** Identity + order
- **Examples:** Likert scales (Strongly Disagree to Strongly Agree), education level, race finishing position, satisfaction ratings
- **Permissible statistics:** Median, percentiles, rank-based (non-parametric) tests
- **Not permissible:** Mean (technically), standard deviation

::: {.callout-warning}
## The Likert Scale Debate
Whether Likert-scale data can be treated as interval (and thus analyzed with means and parametric tests) is a long-standing debate in statistics. Many researchers treat 5- or 7-point Likert scales as approximately interval, particularly when summing multiple items into a composite score. Be prepared to justify your choice.
:::

### Interval Scale

Interval variables have **equal intervals** between consecutive values, but there is **no true zero point** — zero does not mean "none of the quantity."

- **Properties:** Identity + order + equal intervals
- **Examples:** Temperature in Celsius or Fahrenheit, calendar year, IQ scores
- **Permissible statistics:** Mean, standard deviation, correlation, t-tests
- **Not permissible:** Ratios (20°C is not "twice as hot" as 10°C)

### Ratio Scale

Ratio variables possess all properties of interval scales **plus a true zero point**, meaning zero indicates the complete absence of the quantity. Ratios between values are meaningful.

- **Properties:** Identity + order + equal intervals + true zero
- **Examples:** Height, weight, income, reaction time, temperature in Kelvin
- **Permissible statistics:** All statistical operations, including geometric mean and coefficient of variation
- **Note:** This is the most informative scale and the most common in quantitative research

### Summary Table

| Scale | Order | Equal Intervals | True Zero | Examples | Key Statistics |
|-------|-------|----------------|-----------|----------|----------------|
| Nominal | ✗ | ✗ | ✗ | Blood type, species | Mode, frequency |
| Ordinal | ✓ | ✗ | ✗ | Likert, rank | Median, percentile |
| Interval | ✓ | ✓ | ✗ | Temperature (°C), IQ | Mean, SD, correlation |
| Ratio | ✓ | ✓ | ✓ | Height, weight, income | All operations |

: Stevens' Four Levels of Measurement {.striped}

### Continuous vs. Discrete Variables

Independently of scale, variables are also classified as:

- **Discrete:** Take countable values, often integers (number of children, number of visits, count of defects). There are gaps between possible values.
- **Continuous:** Can take any value within a range (height, temperature, time). Between any two values, infinitely many values are possible.

This distinction determines whether we use a **probability mass function** (discrete) or a **probability density function** (continuous) — covered in Sections 4 and 5.

## Example: Identifying Measurement Scales

**Example 2.1.** A health survey collects the following variables. Identify the measurement scale for each.

| Variable | Values Recorded | Scale | Justification |
|----------|----------------|-------|---------------|
| Patient ID | 001, 002, 003 | Nominal | Labels only; no order or arithmetic |
| Pain level | 1 (none) to 10 (severe) | Ordinal | Ordered, but intervals not guaranteed equal |
| Body temperature | 36.5°C, 37.2°C | Interval | Equal intervals; 0°C ≠ absence of temperature |
| Blood pressure | 120 mmHg, 80 mmHg | Ratio | True zero exists; 120 is twice 60 mmHg |
| Hospital ward | Ward A, Ward B | Nominal | Categories with no meaningful order |
| Recovery days | 3, 7, 14 | Ratio | True zero (0 days = no recovery needed) |

: Example 2.1 — Identifying Measurement Scales {.striped}

**Key insight:** Patient ID looks like a number but is nominal. Recovery days looks similar to pain level, but has a true zero. Always ask: *Does zero mean "none of this"?* and *Are the intervals equal?*

## R Example: Identifying and Converting Variable Types

```{r data-types}
# --- Load a dataset and inspect variable types ---
data(mtcars)

# Check R's automatic type assignment
str(mtcars)
```

```{r data-types-2}
# --- Several variables in mtcars are stored as numeric
#     but should be treated as categorical (nominal/ordinal)
#     cyl: number of cylinders (4, 6, 8) — ordinal/nominal
#     am:  transmission type (0=auto, 1=manual) — nominal
#     gear: number of gears — ordinal

# Convert to appropriate types
mtcars_clean <- mtcars |>
  mutate(
    cyl  = factor(cyl,  levels = c(4, 6, 8),
                  labels = c("4-cyl", "6-cyl", "8-cyl"),
                  ordered = TRUE),          # ordinal
    am   = factor(am,   levels = c(0, 1),
                  labels = c("Automatic", "Manual")),  # nominal
    gear = factor(gear, levels = c(3, 4, 5),
                  ordered = TRUE)           # ordinal
  )

# Inspect the corrected structure
str(mtcars_clean)
```

```{r data-types-plot}
# --- Visualize appropriate to scale ---
# Nominal/Ordinal: bar chart (NOT histogram — no meaningful continuity)
p1 <- ggplot(mtcars_clean, aes(x = cyl, fill = cyl)) +
  geom_bar(color = "white") +
  scale_fill_brewer(palette = "Blues") +
  labs(title = "Ordinal Variable: Cylinders",
       subtitle = "Bar chart — appropriate for categorical data",
       x = "Cylinders", y = "Count") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none")

# Ratio: histogram (continuous, true zero)
p2 <- ggplot(mtcars_clean, aes(x = wt)) +
  geom_histogram(bins = 10, fill = "steelblue",
                 color = "white", alpha = 0.8) +
  labs(title = "Ratio Variable: Car Weight",
       subtitle = "Histogram — appropriate for continuous data",
       x = "Weight (1000 lbs)", y = "Count") +
  theme_minimal(base_size = 12)

p1 + p2
```

**Code explanation:**

- `factor(..., ordered = TRUE)` creates an ordered factor in R, correctly representing ordinal data.
- `str()` reveals how R internally stores each variable — a critical first step in any analysis.
- The choice of plot type (bar chart vs. histogram) follows directly from the measurement scale: bar charts for categorical, histograms for continuous.

**Try modifying:** Use `class()` and `levels()` on the converted variables to confirm their types and order.

## Exercises

::: {.callout-tip}
## Exercise 2.1
For each variable below, identify: (a) the measurement scale, (b) whether it is discrete or continuous, and (c) the most appropriate graph type.

1. Number of customer complaints per day
2. Country of birth of survey respondents
3. Annual household income in USD
4. Student satisfaction (1 = Very Dissatisfied, 5 = Very Satisfied)
5. Time taken to complete an online task (in seconds)
6. Zip code of a business location
:::

::: {.callout-tip}
## Exercise 2.2
Load the `iris` dataset in R. For each of the five variables, identify the measurement scale. Convert any variables that are incorrectly stored in R to their appropriate type using `factor()`. Justify each decision.
:::

---

# Probability Fundamentals

## Introduction

Probability is the mathematical framework for **reasoning under uncertainty**. Every time a data scientist makes a prediction, estimates a parameter, or runs a hypothesis test, they are implicitly using probability theory. Understanding the foundational rules — not just the formulas — is essential for building correct intuition about statistical methods.

## Theory

### Sample Space and Events

A **random experiment** is any process whose outcome is uncertain. The **sample space** $S$ is the set of all possible outcomes. An **event** $A$ is any subset of the sample space.

$$S = \{\text{all possible outcomes}\}, \qquad A \subseteq S$$

**Example:** Rolling a six-sided die: $S = \{1, 2, 3, 4, 5, 6\}$. The event "rolling an even number" is $A = \{2, 4, 6\}$.

### Kolmogorov's Axioms

All of probability theory rests on three axioms proposed by Andrey Kolmogorov in 1933:

1. **Non-negativity:** $P(A) \geq 0$ for any event $A$
2. **Normalization:** $P(S) = 1$
3. **Countable additivity:** For mutually exclusive events $A_1, A_2, \ldots$:
$$P(A_1 \cup A_2 \cup \cdots) = P(A_1) + P(A_2) + \cdots$$

From these three axioms, all other probability rules are derived.

### Derived Rules

**Complement Rule:**
$$P(A^c) = 1 - P(A)$$

**Addition Rule** (for any two events):
$$P(A \cup B) = P(A) + P(B) - P(A \cap B)$$

**Multiplication Rule** (for independent events $A \perp B$):
$$P(A \cap B) = P(A) \cdot P(B)$$

**Law of Total Probability:** If $B_1, B_2, \ldots, B_k$ partition the sample space:
$$P(A) = \sum_{i=1}^{k} P(A \mid B_i) \cdot P(B_i)$$

### Frequentist vs. Bayesian Interpretation

Two schools of thought define what a probability *means*:

| Interpretation | Definition | Use Case |
|---------------|------------|----------|
| **Frequentist** | Long-run relative frequency in repeated trials | Classical hypothesis testing |
| **Bayesian** | Degree of belief, updated as evidence arrives | Bayesian inference, machine learning |

: Two interpretations of probability {.striped}

Both interpretations are mathematically consistent with Kolmogorov's axioms — they differ in *meaning*, not in the rules.

## Example: Probability Rules

**Example 2.2.** A data science program has 60 students. From records: 35 have programming experience (P), 28 have statistics experience (S), and 15 have both.

**Find:** (a) $P(P \cup S)$, (b) $P(\text{neither})$, (c) $P(P \text{ only})$

**Step 1 — Convert to probabilities:**
$$P(P) = \frac{35}{60} \approx 0.583, \quad P(S) = \frac{28}{60} \approx 0.467, \quad P(P \cap S) = \frac{15}{60} = 0.25$$

**(a) Addition Rule:**
$$P(P \cup S) = 0.583 + 0.467 - 0.25 = 0.80$$

So 80% of students have at least one type of experience.

**(b) Complement Rule:**
$$P(\text{neither}) = 1 - P(P \cup S) = 1 - 0.80 = 0.20$$

20% of students have neither programming nor statistics experience.

**(c) P only (programming but not statistics):**
$$P(P \cap S^c) = P(P) - P(P \cap S) = 0.583 - 0.25 = 0.333$$

## R Example: Probability Rules and Simulation

```{r prob-simulation}
# --- Simulate probability with repeated experiments ---
# Verify the Addition Rule via simulation

set.seed(123)
n_trials <- 100000

# Simulate 60-student cohorts: each student gets P and S independently
# P(P) = 35/60, P(S) = 28/60, P(P and S) = 15/60
# We'll simulate individual student memberships

p_prog  <- 35 / 60
p_stats <- 28 / 60
p_both  <- 15 / 60

# Simulate events for each trial
has_prog  <- runif(n_trials) < p_prog
has_stats <- runif(n_trials) < p_stats

# Correct for the joint probability (to match P(P∩S) = 15/60)
# Use the conditional: P(S|P) = 15/35
has_stats_given_prog <- runif(n_trials) < (p_both / p_prog)
has_both <- has_prog & has_stats_given_prog
has_stats_not_prog <- (!has_prog) & (runif(n_trials) < ((p_stats - p_both) / (1 - p_prog)))
has_stats_final <- has_both | has_stats_not_prog

# Compute simulated probabilities
sim_results <- data.frame(
  Event       = c("P(Programming)", "P(Statistics)",
                  "P(Both)", "P(At least one)", "P(Neither)"),
  Theoretical = c(35/60, 28/60, 15/60, 0.80, 0.20),
  Simulated   = c(
    mean(has_prog),
    mean(has_stats_final),
    mean(has_both),
    mean(has_prog | has_stats_final),
    mean(!has_prog & !has_stats_final)
  )
)

sim_results[, 2:3] <- round(sim_results[, 2:3], 4)

kable(sim_results,
      caption = "Theoretical vs. Simulated Probabilities (100,000 trials)",
      col.names = c("Event", "Theoretical", "Simulated")) |>
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

```{r prob-venn}
# --- Visualize with a proportional bar chart (Venn alternative) ---
event_df <- data.frame(
  Group       = c("P only", "S only", "Both P & S", "Neither"),
  Count       = c(20, 13, 15, 12),
  Proportion  = c(20/60, 13/60, 15/60, 12/60)
)

ggplot(event_df, aes(x = reorder(Group, -Proportion),
                     y = Proportion, fill = Group)) +
  geom_col(color = "white", width = 0.6) +
  geom_text(aes(label = paste0(round(Proportion * 100, 1), "%")),
            vjust = -0.5, size = 4, fontface = "bold") +
  scale_fill_brewer(palette = "Set2") +
  scale_y_continuous(labels = scales::percent, limits = c(0, 0.45)) +
  labs(title    = "Student Experience Groups",
       subtitle = "Proportional breakdown of programming and statistics experience",
       x        = "Group", y = "Proportion") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none")
```

**Code explanation:**

- `runif(n_trials)` generates uniform random numbers between 0 and 1; comparing to a probability threshold simulates Bernoulli trials.
- `set.seed(123)` ensures reproducibility — always set a seed when using random number generators.
- Simulation confirms theoretical results: with 100,000 trials, simulated and theoretical probabilities converge (Law of Large Numbers).

## Exercises

::: {.callout-tip}
## Exercise 2.3
A quality control inspector examines products from two machines. Machine A produces 60% of output with a 3% defect rate. Machine B produces 40% with a 5% defect rate.

(a) What is the probability that a randomly selected product is defective?  
(b) If a product is not defective, what is the probability it came from Machine A?  
(c) Verify your answers using simulation in R (at least 50,000 trials).
:::

::: {.callout-tip}
## Exercise 2.4
Using the `sample()` function in R, simulate rolling two fair dice 10,000 times.

(a) Estimate $P(\text{sum} = 7)$ and compare to the theoretical value.  
(b) Estimate $P(\text{sum} > 9)$.  
(c) Plot the simulated distribution of sums as a bar chart.
:::

---

# Conditional Probability and Bayes' Theorem

## Introduction

In practice, we rarely ask about probabilities in a vacuum. We ask: *Given that a patient tested positive, what is the probability they actually have the disease?* Or: *Given that a customer clicked an ad, what is the probability they will purchase?* **Conditional probability** formalizes how information about one event changes our assessment of another. **Bayes' Theorem** takes this further, providing a rigorous method for updating beliefs in light of new evidence — the foundation of Bayesian statistics and many machine learning algorithms.

## Theory

### Conditional Probability

The **conditional probability** of event $A$ given that event $B$ has occurred is:

$$P(A \mid B) = \frac{P(A \cap B)}{P(B)}, \qquad \text{provided } P(B) > 0$$

Intuitively, we restrict our universe to the world where $B$ has already happened, and ask what fraction of that restricted world includes $A$.

**Independence** is a special case: $A$ and $B$ are independent if and only if:
$$P(A \mid B) = P(A) \qquad \Longleftrightarrow \qquad P(A \cap B) = P(A) \cdot P(B)$$

### Bayes' Theorem

Bayes' Theorem inverts conditional probabilities — it tells us $P(A \mid B)$ when we know $P(B \mid A)$:

$$P(A \mid B) = \frac{P(B \mid A) \cdot P(A)}{P(B)}$$

Expanding the denominator using the Law of Total Probability:

$$P(A \mid B) = \frac{P(B \mid A) \cdot P(A)}{P(B \mid A) \cdot P(A) + P(B \mid A^c) \cdot P(A^c)}$$

In Bayesian language:

| Term | Name | Meaning |
|------|------|---------|
| $P(A)$ | **Prior** | Belief about $A$ before seeing evidence $B$ |
| $P(B \mid A)$ | **Likelihood** | Probability of the evidence if $A$ is true |
| $P(B)$ | **Marginal likelihood** | Total probability of the evidence |
| $P(A \mid B)$ | **Posterior** | Updated belief after seeing evidence $B$ |

: Components of Bayes' Theorem {.striped}

### Sensitivity, Specificity, and Predictive Value

In medical testing and classification, Bayes' Theorem connects several important metrics:

- **Sensitivity** (True Positive Rate): $P(\text{Test}^+ \mid \text{Disease}^+)$
- **Specificity** (True Negative Rate): $P(\text{Test}^- \mid \text{Disease}^-)$
- **Positive Predictive Value (PPV)**: $P(\text{Disease}^+ \mid \text{Test}^+)$ — computed via Bayes

$$\text{PPV} = \frac{\text{Sensitivity} \times \text{Prevalence}}{\text{Sensitivity} \times \text{Prevalence} + (1 - \text{Specificity}) \times (1 - \text{Prevalence})}$$

## Example: Medical Diagnosis with Bayes' Theorem

**Example 2.3.** A disease affects 1% of the population (prevalence = 0.01). A diagnostic test has:

- Sensitivity = 0.95: $P(\text{Test}^+ \mid \text{Disease}) = 0.95$
- Specificity = 0.90: $P(\text{Test}^- \mid \text{No Disease}) = 0.90$, so $P(\text{Test}^+ \mid \text{No Disease}) = 0.10$

**Question:** If a patient tests positive, what is the probability they actually have the disease?

**Step 1 — Prior:** $P(D) = 0.01$, $P(D^c) = 0.99$

**Step 2 — Marginal probability of a positive test:**
$$P(\text{Test}^+) = P(\text{Test}^+ \mid D) \cdot P(D) + P(\text{Test}^+ \mid D^c) \cdot P(D^c)$$
$$= (0.95)(0.01) + (0.10)(0.99) = 0.0095 + 0.099 = 0.1085$$

**Step 3 — Posterior (PPV):**
$$P(D \mid \text{Test}^+) = \frac{(0.95)(0.01)}{0.1085} = \frac{0.0095}{0.1085} \approx 0.0876$$

**Interpretation:** Despite a positive test, there is only an **8.76% chance** the patient actually has the disease. This counterintuitive result — known as the **base rate fallacy** — occurs because the disease is rare (1%) and the false positive rate (10%) is not negligible. Prevalence (the prior) matters enormously in diagnostic settings.

## R Example: Bayes' Theorem and the Base Rate Effect

```{r bayes}
# --- Bayes' Theorem function ---
bayes_ppv <- function(sensitivity, specificity, prevalence) {
  p_pos_given_d  <- sensitivity
  p_pos_given_nd <- 1 - specificity
  p_d            <- prevalence
  p_nd           <- 1 - prevalence

  p_pos <- p_pos_given_d * p_d + p_pos_given_nd * p_nd
  ppv   <- (p_pos_given_d * p_d) / p_pos
  npv   <- (specificity * p_nd) / (1 - p_pos)

  cat("=== Bayes' Theorem: Diagnostic Test ===\n")
  cat("Sensitivity:       ", sensitivity, "\n")
  cat("Specificity:       ", specificity, "\n")
  cat("Prevalence:        ", prevalence, "\n")
  cat("P(Test+):          ", round(p_pos, 4), "\n")
  cat("PPV (P(D|Test+)):  ", round(ppv, 4), "\n")
  cat("NPV (P(D^c|Test-)): ", round(npv, 4), "\n")
}

# Run for the example
bayes_ppv(sensitivity = 0.95, specificity = 0.90, prevalence = 0.01)
```

```{r bayes-plot}
# --- How does PPV change as prevalence changes? ---
prevalence_seq <- seq(0.001, 0.30, by = 0.001)

ppv_values <- sapply(prevalence_seq, function(prev) {
  p_pos <- 0.95 * prev + 0.10 * (1 - prev)
  (0.95 * prev) / p_pos
})

ppv_df <- data.frame(prevalence = prevalence_seq, ppv = ppv_values)

ggplot(ppv_df, aes(x = prevalence, y = ppv)) +
  geom_line(color = "steelblue", linewidth = 1.3) +
  geom_hline(yintercept = 0.50, linetype = "dashed",
             color = "tomato", linewidth = 0.9) +
  geom_vline(xintercept = 0.01, linetype = "dotted",
             color = "darkgreen", linewidth = 0.9) +
  annotate("text", x = 0.018, y = 0.12,
           label = "Prevalence = 1%\nPPV ≈ 8.8%",
           color = "darkgreen", size = 3.8, hjust = 0) +
  annotate("text", x = 0.18, y = 0.45,
           label = "PPV = 50% threshold",
           color = "tomato", size = 3.8) +
  scale_x_continuous(labels = scales::percent) +
  scale_y_continuous(labels = scales::percent) +
  labs(title    = "Positive Predictive Value vs. Disease Prevalence",
       subtitle = "Sensitivity = 95%, Specificity = 90%",
       x        = "Disease Prevalence",
       y        = "Positive Predictive Value (PPV)") +
  theme_minimal(base_size = 13)
```

**Code explanation:**

- `bayes_ppv()` is a reusable function that takes sensitivity, specificity, and prevalence as arguments — you can apply it to any diagnostic scenario.
- `sapply()` applies the PPV computation across a vector of prevalence values, enabling the sensitivity curve.
- The plot shows a critical insight: PPV improves dramatically as prevalence increases, even with fixed test characteristics.

**Try modifying:** Change specificity to 0.99 and observe how PPV improves at low prevalence. This illustrates why high specificity matters so much for screening rare conditions.

## Exercises

::: {.callout-tip}
## Exercise 2.5
A spam filter has sensitivity = 0.98 (correctly identifies spam) and specificity = 0.95 (correctly identifies legitimate email). In a typical inbox, 20% of emails are spam.

(a) What is the probability that an email flagged as spam is actually spam (PPV)?  
(b) What is the probability that an email passed as legitimate is actually spam (1 - NPV)?  
(c) How do these values change if spam increases to 40% of all emails? Use your R function.
:::

::: {.callout-tip}
## Exercise 2.6 (Challenge)
A factory uses two independent tests to detect a defect (prevalence = 5%). Test 1: sensitivity = 0.90, specificity = 0.85. Test 2: sensitivity = 0.85, specificity = 0.92. A product is flagged only if **both** tests are positive.

(a) Derive the combined sensitivity and specificity of the two-test system.  
(b) Compute the PPV of the combined system.  
(c) Compare to using Test 1 alone. Which system is more reliable?
:::

---

# Discrete Probability Distributions

## Introduction

A **probability distribution** is a mathematical function that assigns probabilities to all possible values of a random variable. For **discrete** random variables — those that take countable values — the distribution is described by a **probability mass function (PMF)**. In data science, discrete distributions model count data, binary outcomes, and rare events. Two distributions are fundamental: the **Binomial** (fixed trials with two outcomes) and the **Poisson** (counts of rare events over an interval).

## Theory

### The Binomial Distribution

The **Binomial distribution** $X \sim \text{Binomial}(n, p)$ models the number of successes in $n$ **independent** trials, each with success probability $p$.

**PMF:**
$$P(X = k) = \binom{n}{k} p^k (1-p)^{n-k}, \qquad k = 0, 1, \ldots, n$$

where $\binom{n}{k} = \frac{n!}{k!(n-k)!}$ is the binomial coefficient (number of ways to choose $k$ from $n$).

**Key properties:**

| Property | Formula |
|----------|---------|
| Mean | $\mu = np$ |
| Variance | $\sigma^2 = np(1-p)$ |
| Standard Deviation | $\sigma = \sqrt{np(1-p)}$ |

: Binomial distribution properties {.striped}

**Assumptions:** Fixed $n$, constant $p$, independent trials, binary outcome (success/failure).

### The Poisson Distribution

The **Poisson distribution** $X \sim \text{Poisson}(\lambda)$ models the number of events occurring in a fixed interval of time or space, when events occur independently at a constant average rate $\lambda$.

**PMF:**
$$P(X = k) = \frac{e^{-\lambda} \lambda^k}{k!}, \qquad k = 0, 1, 2, \ldots$$

**Key properties:**

| Property | Formula |
|----------|---------|
| Mean | $\mu = \lambda$ |
| Variance | $\sigma^2 = \lambda$ |
| Standard Deviation | $\sigma = \sqrt{\lambda}$ |

: Poisson distribution properties {.striped}

Note that in the Poisson distribution, the **mean equals the variance** — a useful diagnostic. If observed count data shows variance much larger than the mean, **overdispersion** is present and a negative binomial model may be more appropriate.

**Poisson as approximation to Binomial:** When $n$ is large and $p$ is small (rare events), Binomial$(n, p) \approx$ Poisson$(\lambda = np)$.

## Example: Binomial and Poisson Distributions

**Example 2.4 — Binomial.** A multiple-choice exam has 20 questions, each with 4 options. A student guesses randomly on every question. Let $X$ = number of correct answers.

$X \sim \text{Binomial}(n = 20, p = 0.25)$

- $\mu = 20 \times 0.25 = 5$ correct answers expected
- $\sigma^2 = 20 \times 0.25 \times 0.75 = 3.75$, $\sigma \approx 1.94$

**Probability of passing (at least 10 correct):**
$$P(X \geq 10) = 1 - P(X \leq 9) = 1 - \sum_{k=0}^{9} \binom{20}{k}(0.25)^k(0.75)^{20-k} \approx 0.014$$

There is only a 1.4% chance of passing by pure guessing.

**Example 2.5 — Poisson.** A hospital emergency room receives an average of 8 patients per hour. Let $X$ = number of patients in the next hour.

$X \sim \text{Poisson}(\lambda = 8)$

- $P(X = 10) = \frac{e^{-8} \cdot 8^{10}}{10!} \approx 0.099$ (about 9.9% chance of exactly 10 arrivals)
- $P(X > 12) = 1 - P(X \leq 12) \approx 0.064$ (6.4% chance of more than 12 arrivals)

## R Example: Binomial and Poisson Distributions

```{r discrete-dist}
# === BINOMIAL DISTRIBUTION ===
# X ~ Binomial(n=20, p=0.25)
n <- 20; p <- 0.25
x_binom <- 0:n

# PMF and CDF
binom_df <- data.frame(
  x    = x_binom,
  pmf  = dbinom(x_binom, size = n, prob = p),
  cdf  = pbinom(x_binom, size = n, prob = p)
)

# Probability of passing (X >= 10)
p_pass <- 1 - pbinom(9, size = n, prob = p)
cat("Binomial(20, 0.25):\n")
cat("  Mean:               ", n * p, "\n")
cat("  Variance:           ", n * p * (1-p), "\n")
cat("  P(X >= 10) [pass]:  ", round(p_pass, 4), "\n\n")

# === POISSON DISTRIBUTION ===
# X ~ Poisson(lambda = 8)
lambda <- 8
x_pois <- 0:25

pois_df <- data.frame(
  x   = x_pois,
  pmf = dpois(x_pois, lambda = lambda),
  cdf = ppois(x_pois, lambda = lambda)
)

p_gt12 <- 1 - ppois(12, lambda = lambda)
cat("Poisson(lambda = 8):\n")
cat("  Mean = Variance:    ", lambda, "\n")
cat("  P(X = 10):          ", round(dpois(10, lambda), 4), "\n")
cat("  P(X > 12):          ", round(p_gt12, 4), "\n")
```

```{r discrete-plot}
# --- Plot both PMFs side by side ---
p1 <- ggplot(binom_df, aes(x = x, y = pmf)) +
  geom_col(fill = "steelblue", color = "white", alpha = 0.85) +
  geom_col(data = subset(binom_df, x >= 10),
           aes(x = x, y = pmf), fill = "tomato", alpha = 0.85) +
  annotate("text", x = 15, y = 0.18,
           label = paste0("P(X≥10) = ", round(p_pass, 3)),
           color = "tomato", fontface = "bold", size = 4) +
  labs(title    = "Binomial(n=20, p=0.25)",
       subtitle = "Red bars: P(pass) = P(X ≥ 10)",
       x        = "Number of Correct Answers",
       y        = "Probability") +
  theme_minimal(base_size = 12)

p2 <- ggplot(pois_df, aes(x = x, y = pmf)) +
  geom_col(fill = "seagreen", color = "white", alpha = 0.85) +
  geom_col(data = subset(pois_df, x > 12),
           aes(x = x, y = pmf), fill = "tomato", alpha = 0.85) +
  annotate("text", x = 18, y = 0.14,
           label = paste0("P(X>12) = ", round(p_gt12, 3)),
           color = "tomato", fontface = "bold", size = 4) +
  labs(title    = "Poisson(λ=8)",
       subtitle = "Red bars: P(X > 12) — surge risk",
       x        = "Number of Patients per Hour",
       y        = "Probability") +
  theme_minimal(base_size = 12)

p1 + p2
```

**Code explanation:**

- `dbinom(x, size, prob)` computes the PMF (exact probability at each value); `pbinom()` computes the CDF (cumulative probability up to and including $x$).
- `dpois(x, lambda)` and `ppois(x, lambda)` are the Poisson equivalents.
- Highlighting specific bars (using `subset()` inside a second `geom_col()`) draws attention to the probability region of interest — a technique useful in reports and teaching.

## Exercises

::: {.callout-tip}
## Exercise 2.7
A new medical treatment is effective in 70% of patients. In a clinical trial of 15 patients:

(a) What is the expected number of patients who respond?  
(b) Find $P(X \geq 12)$ — the probability that at least 12 patients respond.  
(c) Plot the full PMF and shade the region corresponding to $P(X \geq 12)$.  
(d) What sample size would be needed so that $P(X \geq 0.8n) \geq 0.90$? Explore this in R.
:::

::: {.callout-tip}
## Exercise 2.8
Website traffic follows a Poisson process with an average of 5 visits per minute.

(a) Find the probability of exactly 3 visits in one minute.  
(b) Find the probability of more than 8 visits in one minute.  
(c) In a 3-minute window, what distribution does the total visits follow, and what is the probability of more than 20 visits?  
(d) Check whether the mean and variance of a Poisson(5) sample of 1,000 observations are approximately equal. Simulate in R.
:::

---

# Continuous Probability Distributions

## Introduction

When a random variable can take any value within a continuous range, we describe its distribution using a **probability density function (PDF)** $f(x)$. Unlike a PMF, the PDF does not give the probability at a single point (which is always zero for continuous variables) — instead, probabilities are areas under the curve:

$$P(a \leq X \leq b) = \int_a^b f(x)\, dx$$

Three continuous distributions are essential foundations in data science: the **Normal**, the **Uniform**, and the **Exponential**.

## Theory

### The Normal (Gaussian) Distribution

$X \sim N(\mu, \sigma^2)$ is the most important distribution in statistics. Its symmetric, bell-shaped PDF is:

$$f(x) = \frac{1}{\sigma\sqrt{2\pi}} \exp\!\left(-\frac{(x-\mu)^2}{2\sigma^2}\right), \qquad -\infty < x < \infty$$

**The standard normal** $Z \sim N(0, 1)$ is obtained by standardizing: $Z = (X - \mu)/\sigma$.

**Empirical Rule (68-95-99.7 rule):**

$$P(\mu - \sigma \leq X \leq \mu + \sigma) \approx 0.6827$$
$$P(\mu - 2\sigma \leq X \leq \mu + 2\sigma) \approx 0.9545$$
$$P(\mu - 3\sigma \leq X \leq \mu + 3\sigma) \approx 0.9973$$

### The Uniform Distribution

$X \sim \text{Uniform}(a, b)$ assigns equal probability density to every value between $a$ and $b$:

$$f(x) = \frac{1}{b-a}, \quad a \leq x \leq b$$

**Mean:** $\mu = \frac{a+b}{2}$, **Variance:** $\sigma^2 = \frac{(b-a)^2}{12}$

The Uniform distribution is used in random number generation, simulation, and as a non-informative (flat) prior in Bayesian analysis.

### The Exponential Distribution

$X \sim \text{Exponential}(\lambda)$ models the **waiting time** between consecutive Poisson events:

$$f(x) = \lambda e^{-\lambda x}, \quad x \geq 0$$

**Mean:** $\mu = 1/\lambda$, **Variance:** $\sigma^2 = 1/\lambda^2$

The exponential distribution has the **memoryless property**: $P(X > s + t \mid X > s) = P(X > t)$. The future waiting time does not depend on how long you have already waited — relevant for modelling system failures, service times, and inter-arrival times.

## Example: Normal, Uniform, and Exponential

**Example 2.6 — Normal.** IQ scores are distributed as $N(\mu = 100, \sigma^2 = 15^2)$.

**(a)** $P(85 \leq X \leq 115)$ — within one standard deviation:
By the empirical rule $\approx 68.27\%$. Exact: $P(-1 \leq Z \leq 1) = 0.6827$.

**(b)** $P(X > 130)$ — probability of "gifted" range:
$$Z = \frac{130 - 100}{15} = 2.0 \qquad P(X > 130) = P(Z > 2) = 1 - 0.9772 = 0.0228$$

About 2.3% of the population has IQ above 130.

**Example 2.7 — Exponential.** A server processes requests at a rate of $\lambda = 4$ per minute (mean waiting time = 15 seconds = 0.25 minutes).

$P(X > 0.5) = e^{-4 \times 0.5} = e^{-2} \approx 0.135$ — 13.5% chance of waiting more than 30 seconds.

## R Example: Continuous Distributions

```{r continuous-dist}
# === NORMAL DISTRIBUTION ===
mu <- 100; sigma <- 15

# Empirical rule verification
p_1sd <- pnorm(mu + sigma, mu, sigma) - pnorm(mu - sigma, mu, sigma)
p_2sd <- pnorm(mu + 2*sigma, mu, sigma) - pnorm(mu - 2*sigma, mu, sigma)
p_gt130 <- 1 - pnorm(130, mu, sigma)

cat("Normal(mu=100, sigma=15):\n")
cat("  P(within 1 SD): ", round(p_1sd, 4), "\n")
cat("  P(within 2 SD): ", round(p_2sd, 4), "\n")
cat("  P(IQ > 130):    ", round(p_gt130, 4), "\n\n")

# === EXPONENTIAL DISTRIBUTION ===
lambda_exp <- 4  # rate = 4 per minute
p_wait_gt30sec <- 1 - pexp(0.5, rate = lambda_exp)
cat("Exponential(lambda=4 per min):\n")
cat("  Mean wait time (min): ", 1/lambda_exp, "\n")
cat("  P(wait > 30 sec):     ", round(p_wait_gt30sec, 4), "\n")
```

```{r continuous-plot}
# --- Plot all three distributions ---
x_norm <- seq(55, 145, length.out = 300)
x_unif <- seq(-0.2, 1.2, length.out = 300)
x_exp  <- seq(0, 2, length.out = 300)

# Normal
norm_df <- data.frame(x = x_norm, y = dnorm(x_norm, 100, 15),
                      shade = x_norm >= 130)
p1 <- ggplot(norm_df, aes(x, y)) +
  geom_line(color = "steelblue", linewidth = 1.2) +
  geom_area(data = subset(norm_df, shade),
            fill = "tomato", alpha = 0.5) +
  annotate("text", x = 135, y = 0.006,
           label = "P(X>130)\n≈ 2.3%", color = "tomato", size = 3.5) +
  labs(title = "Normal(μ=100, σ=15)",
       x = "IQ Score", y = "Density") +
  theme_minimal(base_size = 11)

# Uniform
unif_df <- data.frame(x = x_unif,
                      y = dunif(x_unif, min = 0, max = 1))
p2 <- ggplot(unif_df, aes(x, y)) +
  geom_line(color = "seagreen", linewidth = 1.2) +
  geom_area(data = subset(unif_df, x >= 0 & x <= 1),
            fill = "seagreen", alpha = 0.3) +
  labs(title = "Uniform(0, 1)",
       x = "Value", y = "Density") +
  theme_minimal(base_size = 11)

# Exponential
exp_df <- data.frame(x = x_exp, y = dexp(x_exp, rate = 4),
                     shade = x_exp > 0.5)
p3 <- ggplot(exp_df, aes(x, y)) +
  geom_line(color = "darkorange", linewidth = 1.2) +
  geom_area(data = subset(exp_df, shade),
            fill = "tomato", alpha = 0.5) +
  annotate("text", x = 1, y = 1.2,
           label = "P(X>0.5)\n≈ 13.5%", color = "tomato", size = 3.5) +
  labs(title = "Exponential(λ=4)",
       x = "Wait Time (min)", y = "Density") +
  theme_minimal(base_size = 11)

p1 + p2 + p3
```

**Code explanation:**

- `dnorm()`, `dunif()`, `dexp()` compute the PDF (density) — used for plotting the curve.
- `pnorm()`, `punif()`, `pexp()` compute the CDF (cumulative probability) — used for probability calculations.
- `qnorm()`, `qunif()`, `qexp()` compute quantiles (inverse CDF) — used to find critical values.
- Shaded areas under the PDF represent probabilities — a key visual concept for students learning integration-based probability.

## Exercises

::: {.callout-tip}
## Exercise 2.9
Heights of adult males in a population follow $N(\mu = 175\text{ cm}, \sigma^2 = 6.25^2)$.

(a) What proportion of males are taller than 185 cm?  
(b) What height marks the 90th percentile?  
(c) Between what two heights (symmetric around the mean) do the middle 95% of males fall?  
(d) Plot the full PDF and shade the region corresponding to (a).
:::

::: {.callout-tip}
## Exercise 2.10
A call center receives calls at a rate of 12 per hour (Poisson process).

(a) What is the distribution of waiting time between calls? State the parameter(s).  
(b) What is the probability of waiting more than 10 minutes between calls?  
(c) Simulate 10,000 inter-arrival times in R and compare the sample mean and variance to theoretical values.  
(d) Plot a histogram of the simulated times with the theoretical PDF overlaid.
:::

---

# Sampling Distributions and the Central Limit Theorem

## Introduction

In practice, we observe a **sample** and wish to draw conclusions about the **population**. The key bridge between sample and population is the concept of a **sampling distribution** — the probability distribution of a statistic (like the sample mean) computed from all possible samples of size $n$. Understanding sampling distributions, and especially the **Central Limit Theorem (CLT)**, is what makes inference possible and explains why the normal distribution appears so ubiquitously in statistics.

## Theory

### Sampling Distribution of the Mean

Suppose we draw a random sample $X_1, X_2, \ldots, X_n$ from a population with mean $\mu$ and variance $\sigma^2$. The sample mean is:
$$\bar{X} = \frac{1}{n}\sum_{i=1}^n X_i$$

The sampling distribution of $\bar{X}$ has:

$$E[\bar{X}] = \mu \qquad \text{(unbiased)}$$
$$\text{Var}(\bar{X}) = \frac{\sigma^2}{n} \qquad \text{(variance decreases with sample size)}$$
$$\text{SE}(\bar{X}) = \frac{\sigma}{\sqrt{n}} \qquad \text{(standard error)}$$

### The Central Limit Theorem (CLT)

::: {.callout-important}
## Central Limit Theorem
Let $X_1, X_2, \ldots, X_n$ be independent and identically distributed random variables from **any distribution** with mean $\mu$ and finite variance $\sigma^2$. Then as $n \to \infty$:

$$\frac{\bar{X} - \mu}{\sigma / \sqrt{n}} \xrightarrow{d} N(0, 1)$$

The standardized sample mean converges in distribution to the standard normal, **regardless of the shape of the original population distribution**.
:::

**Practical rule of thumb:** For most distributions, $n \geq 30$ is sufficient for the normal approximation to be reliable. For highly skewed distributions, larger $n$ may be needed.

**Why this matters:** The CLT justifies the use of $z$-tests and $t$-tests even when the population is not normal, provided the sample is large enough. It explains why the normal distribution appears so frequently in real data — many measurements are averages or sums of numerous independent contributions.

### The $t$-Distribution

When $\sigma$ is unknown (always in practice) and we substitute the sample standard deviation $s$, the standardized mean follows a **Student's $t$-distribution** with $n-1$ degrees of freedom:

$$T = \frac{\bar{X} - \mu}{s / \sqrt{n}} \sim t(n-1)$$

The $t$-distribution is symmetric and bell-shaped like the normal, but has **heavier tails** — reflecting the additional uncertainty from estimating $\sigma$. As $n \to \infty$, $t(n-1) \to N(0,1)$.

## Example: Demonstrating the CLT

**Example 2.8.** Consider sampling from a highly skewed **Exponential(1)** distribution ($\mu = 1$, $\sigma = 1$). The CLT predicts that for large $n$, the sampling distribution of $\bar{X}$ will be approximately $N(1, 1/n)$.

We verify this empirically by taking 10,000 samples of size $n = 2, 5, 30, 100$ and examining the distribution of sample means. As $n$ grows, the histogram of $\bar{X}$ becomes increasingly bell-shaped, regardless of the skewed parent distribution.

## R Example: Demonstrating the Central Limit Theorem

```{r clt}
# --- CLT Demonstration ---
# Parent distribution: Exponential(rate=1) — highly right-skewed
set.seed(42)
n_samples <- 10000
sample_sizes <- c(2, 5, 30, 100)

# Function: draw n_samples means of size n
get_means <- function(n, n_samp = 10000) {
  replicate(n_samp, mean(rexp(n, rate = 1)))
}

# Collect sample means for each sample size
clt_df <- map_dfr(sample_sizes, function(n) {
  data.frame(
    sample_size = paste0("n = ", n),
    xbar        = get_means(n)
  )
})

clt_df$sample_size <- factor(clt_df$sample_size,
                             levels = paste0("n = ", sample_sizes))

# --- Plot ---
ggplot(clt_df, aes(x = xbar)) +
  geom_histogram(aes(y = after_stat(density)),
                 bins = 50, fill = "steelblue",
                 color = "white", alpha = 0.75) +
  stat_function(fun   = dnorm,
                args  = list(mean = 1, sd = 1 / sqrt(as.numeric(
                  gsub("n = ", "", clt_df$sample_size[1])))),
                color = "tomato", linewidth = 1) +
  facet_wrap(~sample_size, scales = "free", ncol = 2) +
  labs(title    = "Central Limit Theorem: Sampling from Exponential(1)",
       subtitle = "As n increases, the distribution of X̄ approaches Normal",
       x        = "Sample Mean (X̄)", y = "Density") +
  theme_minimal(base_size = 12)
```

```{r clt-overlay}
# --- Overlay theoretical normal for each panel ---
clt_summary <- clt_df |>
  mutate(n_val = as.numeric(gsub("n = ", "", sample_size))) |>
  group_by(sample_size, n_val) |>
  summarise(.groups = "drop")

# Better approach: compute theoretical curves per panel
norm_curves <- map_dfr(sample_sizes, function(n) {
  x_seq <- seq(0, 3, length.out = 300)
  data.frame(
    sample_size = paste0("n = ", n),
    x           = x_seq,
    y           = dnorm(x_seq, mean = 1, sd = 1 / sqrt(n))
  )
})
norm_curves$sample_size <- factor(norm_curves$sample_size,
                                  levels = paste0("n = ", sample_sizes))

ggplot(clt_df, aes(x = xbar)) +
  geom_histogram(aes(y = after_stat(density)),
                 bins = 60, fill = "steelblue",
                 color = "white", alpha = 0.7) +
  geom_line(data = norm_curves, aes(x = x, y = y),
            color = "tomato", linewidth = 1.2) +
  facet_wrap(~sample_size, scales = "free", ncol = 2) +
  labs(title    = "CLT: Simulated Sampling Distributions vs. Theoretical Normal",
       subtitle = "Red curve = theoretical N(μ, σ²/n) — fits better as n grows",
       x        = "Sample Mean (X̄)", y        = "Density") +
  theme_minimal(base_size = 12)
```

**Code explanation:**

- `replicate(n_samp, mean(rexp(n, rate=1)))` is the core simulation: repeat `n_samp` times, drawing a sample of size `n` from an Exponential distribution and computing its mean.
- `map_dfr()` from `purrr` (loaded via `tidyverse`) applies a function across a vector and row-binds the results into a data frame — a clean alternative to a `for` loop.
- Separate `norm_curves` data frame computes the theoretical normal PDF for each panel, enabling correct overlay across facets with different scales.

## Exercises

::: {.callout-tip}
## Exercise 2.11
(a) Simulate the CLT using a **Uniform(0, 1)** parent distribution. Draw 10,000 samples of sizes $n = 1, 5, 20, 50$ and plot the sampling distribution of $\bar{X}$ for each.  
(b) At what sample size does the sampling distribution appear approximately normal?  
(c) Compare the simulated standard error to the theoretical value $\sigma/\sqrt{n}$ for each $n$.
:::

::: {.callout-tip}
## Exercise 2.12
A population of exam scores has $\mu = 72$ and $\sigma = 14$ (distribution unknown).

(a) A random sample of $n = 49$ students is drawn. What is the probability that the sample mean exceeds 75?  
(b) How large must $n$ be so that $P(\bar{X} > 74) < 0.05$?  
(c) Why can we use the normal distribution to answer (a) even though the population distribution is unknown?
:::

---

# Assessing Normality

## Introduction

Many statistical tests — including $t$-tests, ANOVA, linear regression residual diagnostics, and Pearson correlation — assume that the data (or the residuals) are approximately normally distributed. **Assessing normality** is therefore a critical step in any data analysis workflow, not a formality. In this section, we cover both graphical methods (visual, flexible) and formal statistical tests (objective, sample-size dependent), and discuss how to interpret their results in practice.

## Theory

### Graphical Methods

**Histogram with normal curve overlay:** A histogram compared to the shape of a normal distribution provides an intuitive first check. Departures such as skewness, bimodality, or heavy tails are immediately visible.

**QQ Plot (Quantile-Quantile Plot):** A QQ plot compares the **quantiles of the observed data** to the **quantiles expected from a normal distribution**. If the data are normal, points fall approximately on a straight diagonal line. Specific patterns indicate specific departures:

| Pattern in QQ Plot | Interpretation |
|-------------------|----------------|
| Points on the line | Approximately normal |
| S-shaped curve | Skewness (left or right) |
| Points curve up at both ends | Heavy tails (leptokurtic) |
| Points curve down at both ends | Light tails (platykurtic) |
| Step pattern | Discrete data |

: Interpreting QQ plot patterns {.striped}

### Formal Normality Tests

Formal tests assess $H_0$: the data follow a normal distribution versus $H_1$: the data do not.

**Shapiro-Wilk Test** (`shapiro.test` in R):
- Best for small to moderate samples ($n \leq 2000$)
- Generally the most powerful normality test
- $W$ statistic close to 1 indicates normality; small $p$-value → reject normality

**Kolmogorov-Smirnov Test** (`ks.test` in R):
- Compares empirical CDF to theoretical CDF
- Less powerful than Shapiro-Wilk; requires estimating parameters separately
- More appropriate for large samples

**Anderson-Darling Test** (`ad.test` in `nortest` package):
- Places more weight on the tails than the KS test
- Recommended when tail behavior is particularly important

::: {.callout-warning}
## A Critical Caveat About Normality Tests
With **large samples** ($n > 200$), normality tests become so powerful that they reject $H_0$ for trivial, practically unimportant deviations from normality. With **small samples**, tests have low power and may fail to detect real non-normality. **Always combine graphical inspection with formal tests**, and use subject-matter knowledge to judge whether any detected departure from normality matters for the analysis.
:::

### Transformations for Non-Normal Data

When normality is violated, transformations can make data approximately normal:

| Pattern | Recommended Transformation |
|---------|---------------------------|
| Right skew | Log: $\log(x)$ or Square root: $\sqrt{x}$ |
| Left skew | Square: $x^2$ or Exponential: $e^x$ |
| Proportions (0 to 1) | Logit: $\log(p/(1-p))$ |
| Count data | Square root: $\sqrt{x}$ |

: Common variance-stabilizing transformations {.striped}

## Example: Assessing Normality of Exam Scores

**Example 2.9.** A researcher records exam scores for 50 students. Summary: $\bar{x} = 74.2$, $s = 11.8$, skewness = $-0.42$. The Shapiro-Wilk test gives $W = 0.973$, $p = 0.32$.

**Interpretation:**

- Skewness of $-0.42$ suggests mild left skew (most students score above the mean, with a few very low scores).
- The QQ plot shows minor departures at the lower tail.
- Shapiro-Wilk $p = 0.32 > 0.05$: we **fail to reject** $H_0$ — no statistically significant evidence against normality.
- **Conclusion:** The data are approximately normal; parametric methods are appropriate.

## R Example: Testing Normality

```{r normality}
# --- Use the 'airquality' built-in dataset ---
# Focus on Ozone concentration (known to be right-skewed)
data(airquality)
ozone <- na.omit(airquality$Ozone)

cat("=== Ozone Concentration — Normality Assessment ===\n")
cat("n:          ", length(ozone), "\n")
cat("Mean:       ", round(mean(ozone), 2), "\n")
cat("Median:     ", median(ozone), "\n")
cat("Skewness:   ", round(moments::skewness(ozone), 3), "\n\n")

# Shapiro-Wilk Test
sw_test <- shapiro.test(ozone)
cat("Shapiro-Wilk Test:\n")
cat("  W =", round(sw_test$statistic, 4),
    "  p-value =", round(sw_test$p.value, 4), "\n\n")

# Anderson-Darling Test (from nortest package)
ad_result <- nortest::ad.test(ozone)
cat("Anderson-Darling Test:\n")
cat("  A =", round(ad_result$statistic, 4),
    "  p-value =", round(ad_result$p.value, 4), "\n")
```

```{r normality-plot}
# --- Four-panel normality assessment ---
ozone_df <- data.frame(ozone = ozone, log_ozone = log(ozone))

p1 <- ggplot(ozone_df, aes(x = ozone)) +
  geom_histogram(aes(y = after_stat(density)), bins = 20,
                 fill = "steelblue", color = "white", alpha = 0.8) +
  stat_function(fun  = dnorm,
                args = list(mean = mean(ozone), sd = sd(ozone)),
                color = "tomato", linewidth = 1.2) +
  labs(title = "A. Histogram: Ozone (raw)",
       x = "Ozone (ppb)", y = "Density") +
  theme_minimal(base_size = 11)

p2 <- ggplot(ozone_df, aes(sample = ozone)) +
  stat_qq(color = "steelblue", size = 1.8) +
  stat_qq_line(color = "tomato", linewidth = 1) +
  labs(title = "B. QQ Plot: Ozone (raw)",
       x = "Theoretical Quantiles", y = "Sample Quantiles") +
  theme_minimal(base_size = 11)

p3 <- ggplot(ozone_df, aes(x = log_ozone)) +
  geom_histogram(aes(y = after_stat(density)), bins = 20,
                 fill = "seagreen", color = "white", alpha = 0.8) +
  stat_function(fun  = dnorm,
                args = list(mean = mean(log(ozone)), sd = sd(log(ozone))),
                color = "tomato", linewidth = 1.2) +
  labs(title = "C. Histogram: log(Ozone)",
       x = "log(Ozone)", y = "Density") +
  theme_minimal(base_size = 11)

p4 <- ggplot(ozone_df, aes(sample = log_ozone)) +
  stat_qq(color = "seagreen", size = 1.8) +
  stat_qq_line(color = "tomato", linewidth = 1) +
  labs(title = "D. QQ Plot: log(Ozone)",
       x = "Theoretical Quantiles", y = "Sample Quantiles") +
  theme_minimal(base_size = 11)

(p1 + p2) / (p3 + p4)
```

```{r normality-transform}
# --- Confirm transformation improves normality ---
sw_log <- shapiro.test(log(ozone))

cat("After log transformation:\n")
cat("  Skewness:  ", round(moments::skewness(log(ozone)), 3), "\n")
cat("  Shapiro-Wilk W =", round(sw_log$statistic, 4),
    "  p-value =", round(sw_log$p.value, 4), "\n")
cat("  Interpretation: p > 0.05 → fail to reject normality ✓\n")
```

**Code explanation:**

- `na.omit()` removes missing values before analysis — always handle `NA`s explicitly.
- `shapiro.test()` is built into base R; `nortest::ad.test()` requires the `nortest` package.
- The four-panel layout (raw vs. log-transformed, histogram vs. QQ) is the standard reporting format for normality assessment — it shows both the problem and the solution.
- `stat_function(fun = dnorm, args = list(...))` overlays the theoretical normal curve fitted to the data's mean and SD.

**Try modifying:** Apply the same normality assessment to the `wind` variable in `airquality`. Is it more or less normal than `ozone`?

## Exercises

::: {.callout-tip}
## Exercise 2.13
Using the `mtcars` dataset, assess the normality of the variables `mpg`, `hp`, and `wt`.

(a) For each variable, produce a histogram with normal overlay and a QQ plot.  
(b) Conduct Shapiro-Wilk and Anderson-Darling tests for each variable.  
(c) Based on your combined assessment, which variable(s) are approximately normal? Which are not?  
(d) For any non-normal variable, apply an appropriate transformation and re-assess.
:::

::: {.callout-tip}
## Exercise 2.14 (Challenge)
Generate three samples of size $n = 20$, $n = 100$, and $n = 500$ from a $t$-distribution with 3 degrees of freedom.

(a) Perform Shapiro-Wilk tests on all three samples. How do $p$-values change with $n$?  
(b) Produce QQ plots for all three. At which sample size does non-normality become visually obvious?  
(c) What does this exercise illustrate about the relationship between sample size and the power of normality tests?
:::

---

# Chapter Lab Activity: Exploring Distributions with the `airquality` Dataset

## Objectives

In this lab, you will apply the full spectrum of Chapter 2 concepts to the `airquality` dataset — daily air quality measurements in New York from May to September 1973. The goal is to characterize each variable's distribution, apply appropriate probability models, and assess normality — building the habit of distribution-first thinking before any analysis.

## The Dataset

```{r lab-intro}
data(airquality)

kable(head(airquality),
      caption = "First 6 Rows of the airquality Dataset") |>
  kable_styling(bootstrap_options = c("striped", "hover"), font_size = 11)

cat("Dimensions:", nrow(airquality), "rows x", ncol(airquality), "columns\n")
cat("Missing values per variable:\n")
print(colSums(is.na(airquality)))
```

**Variable descriptions:**

| Variable | Description | Scale | Type |
|----------|-------------|-------|------|
| `Ozone` | Ozone concentration (ppb) | Ratio | Continuous |
| `Solar.R` | Solar radiation (Langley) | Ratio | Continuous |
| `Wind` | Wind speed (mph) | Ratio | Continuous |
| `Temp` | Temperature (°F) | Interval | Continuous |
| `Month` | Month (5–9) | Ordinal | Discrete |
| `Day` | Day of month (1–31) | Ordinal | Discrete |

## Lab Task 1: Scale Identification and Missing Data

```{r lab-task1}
# Handle missing values and check variable types
airquality_clean <- airquality |>
  mutate(
    Month = factor(Month, levels = 5:9,
                   labels = c("May","Jun","Jul","Aug","Sep"),
                   ordered = TRUE),
    Day   = as.integer(Day)
  )

# Summary of missing data
missing_summary <- data.frame(
  Variable  = names(airquality),
  Type      = c("Ratio","Ratio","Ratio","Interval","Ordinal","Ordinal"),
  N_Missing = colSums(is.na(airquality)),
  Pct_Missing = round(colMeans(is.na(airquality)) * 100, 1)
)
rownames(missing_summary) <- NULL

kable(missing_summary,
      caption = "Variable Types and Missing Data Summary") |>
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE) |>
  column_spec(4, bold = TRUE,
              color = ifelse(missing_summary$Pct_Missing > 0,
                             "tomato", "darkgreen"))
```

## Lab Task 2: Distribution Characterization

```{r lab-task2, fig.height=9}
# Complete distributional profile for all four continuous variables
cont_vars <- c("Ozone", "Solar.R", "Wind", "Temp")
colors    <- c("steelblue", "seagreen", "darkorange", "mediumpurple")

plots <- map2(cont_vars, colors, function(var, col) {
  x <- na.omit(airquality[[var]])
  sw_p <- round(shapiro.test(x)$p.value, 4)
  sk   <- round(moments::skewness(x), 2)

  ggplot(data.frame(x = x), aes(x = x)) +
    geom_histogram(aes(y = after_stat(density)), bins = 20,
                   fill = col, color = "white", alpha = 0.75) +
    geom_density(color = "black", linewidth = 0.9, linetype = "dashed") +
    stat_function(fun  = dnorm,
                  args = list(mean = mean(x), sd = sd(x)),
                  color = "tomato", linewidth = 1) +
    labs(title    = var,
         subtitle = paste0("Skewness = ", sk,
                           " | Shapiro-Wilk p = ", sw_p),
         x        = var, y = "Density") +
    theme_minimal(base_size = 11)
})

(plots[[1]] + plots[[2]]) / (plots[[3]] + plots[[4]])
```

## Lab Task 3: Monthly Distributions and Probability Questions

```{r lab-task3}
# Temperature statistics by month
airquality_clean |>
  group_by(Month) |>
  summarise(
    n      = n(),
    Mean   = round(mean(Temp), 1),
    Median = round(median(Temp), 1),
    SD     = round(sd(Temp), 1),
    Min    = min(Temp),
    Max    = max(Temp)
  ) |>
  kable(caption = "Temperature (°F) Summary by Month") |>
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

```{r lab-task3b}
# Using July temperature as approximately Normal: compute probabilities
july_temp <- na.omit(airquality$Temp[airquality$Month == 7])
mu_jul <- mean(july_temp)
sd_jul <- sd(july_temp)

cat("July Temperature: N(μ =", round(mu_jul,1),
    ", σ =", round(sd_jul,1), ")\n")
cat("P(Temp > 90°F) in July: ",
    round(1 - pnorm(90, mu_jul, sd_jul), 4), "\n")
cat("P(Temp < 80°F) in July: ",
    round(pnorm(80, mu_jul, sd_jul), 4), "\n")
cat("90th percentile of July temp: ",
    round(qnorm(0.90, mu_jul, sd_jul), 1), "°F\n")
```

## Lab Task 4: Normality Assessment and Transformations

```{r lab-task4}
# Systematic normality testing
ozone_clean <- na.omit(airquality$Ozone)

normality_results <- data.frame(
  Variable      = c("Ozone (raw)", "log(Ozone)", "Wind (raw)", "Temp (raw)"),
  Skewness      = round(c(
    moments::skewness(ozone_clean),
    moments::skewness(log(ozone_clean)),
    moments::skewness(na.omit(airquality$Wind)),
    moments::skewness(na.omit(airquality$Temp))
  ), 3),
  SW_statistic  = round(c(
    shapiro.test(ozone_clean)$statistic,
    shapiro.test(log(ozone_clean))$statistic,
    shapiro.test(na.omit(airquality$Wind))$statistic,
    shapiro.test(na.omit(airquality$Temp))$statistic
  ), 4),
  SW_pvalue     = round(c(
    shapiro.test(ozone_clean)$p.value,
    shapiro.test(log(ozone_clean))$p.value,
    shapiro.test(na.omit(airquality$Wind))$p.value,
    shapiro.test(na.omit(airquality$Temp))$p.value
  ), 4),
  Conclusion    = c("Non-normal ✗", "Approx. normal ✓",
                    "Check QQ plot", "Approx. normal ✓")
)

kable(normality_results,
      caption = "Normality Assessment — airquality Variables",
      col.names = c("Variable", "Skewness", "SW Statistic",
                    "SW p-value", "Conclusion")) |>
  kable_styling(bootstrap_options = c("striped", "hover")) |>
  column_spec(5,
              color = ifelse(grepl("✓", normality_results$Conclusion),
                             "darkgreen", "tomato"),
              bold  = TRUE)
```

## Lab Discussion Questions

Answer the following in writing:

1. **Measurement Scales:** The `Month` variable is stored as an integer in R. Explain why it should be converted to an ordered factor for analysis. What statistical operations become inappropriate if we treat it as a ratio-scale number?

2. **Missing Data:** `Ozone` has 24 missing values (15.7%). How might this affect your probability calculations? What assumptions are you implicitly making when you use `na.omit()`?

3. **Distribution Choice:** Based on Lab Task 2, which probability distribution (Normal, Poisson, Exponential, or other) would you use to model each of the four continuous variables? Justify each choice using skewness, shape, and domain knowledge.

4. **Bayes in Context:** Suppose that historically, 30% of summer days in New York have ozone levels considered "unhealthy" (above 100 ppb). A monitoring sensor flags a day as potentially unhealthy with 80% accuracy (sensitivity) and has a 10% false positive rate. Use Bayes' Theorem to compute the probability that a flagged day is truly unhealthy.

5. **CLT Application:** The sample size for each month is small (roughly 30 days). Is the CLT applicable for computing probabilities about monthly mean temperatures? What assumptions are you relying on?

---

# Chapter Summary

## Summary

This chapter established the probabilistic foundation for statistical inference and data science:

- **Measurement scales** (nominal, ordinal, interval, ratio) determine what mathematical operations are permissible and what statistical methods are appropriate. Misidentifying scale leads to meaningless analyses.
- **Probability rules** (complement, addition, multiplication, total probability) provide the logical framework for computing uncertainty. Frequentist and Bayesian interpretations differ in meaning, not in rules.
- **Bayes' Theorem** updates prior beliefs with evidence to yield posterior probabilities — the foundation of diagnostic reasoning, spam filtering, and Bayesian machine learning.
- **Discrete distributions** (Binomial, Poisson) model count and binary outcome data, governed by probability mass functions.
- **Continuous distributions** (Normal, Uniform, Exponential) model measurable quantities via probability density functions and areas under curves.
- **The Central Limit Theorem** is the linchpin of inference: sample means from any distribution converge to normality as $n$ grows, justifying the near-universal use of normal-based methods.
- **Normality assessment** combines graphical inspection (histogram, QQ plot) with formal tests (Shapiro-Wilk, Anderson-Darling). Transformations (log, square root) can correct non-normality.

::: {.callout-important}
## Key Formulas to Know

**Conditional Probability:**
$$P(A \mid B) = \frac{P(A \cap B)}{P(B)}$$

**Bayes' Theorem:**
$$P(A \mid B) = \frac{P(B \mid A) \cdot P(A)}{P(B)}$$

**Binomial PMF:**
$$P(X = k) = \binom{n}{k} p^k (1-p)^{n-k}$$

**Poisson PMF:**
$$P(X = k) = \frac{e^{-\lambda}\lambda^k}{k!}$$

**Standard Error:**
$$\text{SE}(\bar{X}) = \frac{\sigma}{\sqrt{n}}$$

**Standardized Mean (CLT):**
$$Z = \frac{\bar{X} - \mu}{\sigma / \sqrt{n}} \xrightarrow{d} N(0,1)$$
:::

## References

Anderson, E. (1935). The irises of the Gaspé Peninsula. *Bulletin of the American Iris Society*, *59*, 2–5.

Anderson, T. W., & Darling, D. A. (1954). A test of goodness of fit. *Journal of the American Statistical Association*, *49*(268), 765–769.

Box, G. E. P., & Cox, D. R. (1964). An analysis of transformations. *Journal of the Royal Statistical Society: Series B*, *26*(2), 211–243.

Gross, J., & Ligges, U. (2015). *nortest: Tests for normality* (R package version 1.0-4). https://CRAN.R-project.org/package=nortest

Henderson, H. V., & Velleman, P. F. (1981). Building multiple regression models interactively. *Biometrics*, *37*(2), 391–411.

Kolmogorov, A. N. (1933). *Grundbegriffe der Wahrscheinlichkeitsrechnung*. Springer.

Komsta, L., & Novomestky, F. (2022). *moments: Moments, cumulants, skewness, kurtosis and related tests* (R package version 0.14.1). https://CRAN.R-project.org/package=moments

New York State Department of Conservation. (1973). *Daily air quality measurements in New York, May–September 1973*.

Pedersen, T. L. (2022). *patchwork: The composer of plots* (R package version 1.1.2). https://CRAN.R-project.org/package=patchwork

R Core Team. (2024). *R: A language and environment for statistical computing*. R Foundation for Statistical Computing. https://www.R-project.org/

Shapiro, S. S., & Wilk, M. B. (1965). An analysis of variance test for normality. *Biometrika*, *52*(3/4), 591–611.

Stevens, S. S. (1946). On the theory of scales of measurement. *Science*, *103*(2684), 677–680.

Wickham, H. (2016). *ggplot2: Elegant graphics for data analysis*. Springer.

Wickham, H., Averick, M., Bryan, J., Chang, W., McGowan, L. D., François, R., Grolemund, G., Hayes, A., Henry, L., Hester, J., Kuhn, M., Pedersen, T. L., Miller, E., Bache, S. M., Müller, K., Ooms, J., Robinson, D., Seidel, D. P., Spinu, V., … Yutani, H. (2019). Welcome to the tidyverse. *Journal of Open Source Software*, *4*(43), 1686.

Xie, Y. (2015). *Dynamic documents with R and knitr* (2nd ed.). CRC Press.

Yeo, I. K., & Johnson, R. A. (2000). A new family of power transformations to improve normality or symmetry. *Biometrika*, *87*(4), 954–959.

---

*End of Chapter 2. Proceed to Chapter 3: Hypothesis Testing.*
