# Chapter 4: Test of Independence of Variables

```{r setup}
# Load all libraries needed for this chapter
library(tidyverse)
library(knitr)
library(kableExtra)
library(patchwork)
library(vcd)          # association measures, mosaic plots
library(corrplot)     # correlation matrix visualization
library(ggcorrplot)   # ggplot2-based correlation plots
library(rstatix)      # tidy correlation tests
library(ppcor)        # partial correlation
```

---

# Chapter Overview

## Introduction

One of the most fundamental questions in data analysis is whether two variables are **related** to each other, or whether they vary **independently**. Does a patient's smoking status relate to their disease outcome? Is a student's program of study associated with their level of satisfaction? Do income and education move together? Tests of independence answer these questions formally, across all combinations of variable types.

This chapter covers:

- **The Concept of Independence** — theoretical foundations linking back to probability
- **Chi-Square Test of Independence** — testing association between two categorical variables
- **Fisher's Exact Test** — a robust alternative for small or sparse contingency tables
- **Cramér's V and Effect Size** — quantifying the strength of categorical association
- **Correlation Tests** — Pearson and Spearman for continuous and ordinal variables
- **Point-Biserial and Phi Coefficients** — association involving binary variables
- **Partial Correlation** — measuring association while controlling for a third variable

::: {.callout-note}
## Learning Objectives
By the end of this chapter, you will be able to:

1. Distinguish statistical independence from dependence for different variable types.
2. Conduct and interpret chi-square and Fisher's exact tests for categorical data.
3. Compute and interpret Cramér's V as a measure of association strength.
4. Test the significance of Pearson and Spearman correlations.
5. Select the appropriate association measure based on variable types.
6. Compute partial correlations to control for confounding variables.
:::

---

# The Concept of Independence

## Introduction

In Chapter 2, we defined statistical independence in probability terms: two events $A$ and $B$ are independent if $P(A \cap B) = P(A) \cdot P(B)$. In data analysis, we extend this idea to variables: two variables are **statistically independent** if knowing the value of one provides no information about the value of the other. When they are not independent, we say they are **associated** or **dependent**.

Testing for independence is not the same as proving causation. Association tells us that two variables move together — it says nothing about which causes which, or whether both are driven by a third variable. This distinction is critical in data science, where spurious correlations are common in large datasets.

## Theory

### Independence for Categorical Variables

Two categorical variables $X$ and $Y$ are independent if, for all categories $i$ and $j$:

$$P(X = i \cap Y = j) = P(X = i) \cdot P(Y = j)$$

In a contingency table, this means every **observed cell frequency** should equal the corresponding **expected cell frequency** under independence:

$$E_{ij} = \frac{(\text{Row } i \text{ total}) \times (\text{Column } j \text{ total})}{\text{Grand total}} = \frac{R_i \cdot C_j}{N}$$

Any systematic departure of observed frequencies from expected frequencies is evidence of association.

### Independence for Continuous Variables

Two continuous variables $X$ and $Y$ are independent if their **joint distribution** equals the product of their marginal distributions:

$$f(x, y) = f_X(x) \cdot f_Y(y)$$

In practice, we test a weaker condition: **linear independence** (zero correlation). Note that zero correlation does not guarantee full independence — two variables can be nonlinearly related yet have zero Pearson correlation. This is why visualization always accompanies formal tests.

### A Taxonomy of Association Measures

The correct test and measure of association depend on the **types of variables** being compared:

| Variable 1 | Variable 2 | Test | Effect Size |
|-----------|-----------|------|------------|
| Categorical | Categorical | Chi-square / Fisher's | Cramér's V |
| Continuous | Continuous | Pearson / Spearman correlation test | $r$ / $\rho$ |
| Binary | Continuous | Point-biserial correlation test | $r_{pb}$ |
| Binary | Binary | Chi-square / Fisher's | Phi ($\phi$) |
| Continuous (controlling for $Z$) | Continuous | Partial correlation | Partial $r$ |

: Association measures by variable type {.striped}

## Example: Identifying the Correct Test

**Example 4.1.** A researcher collects data on university students. Identify the appropriate test for each pair of variables:

| Variable 1 | Variable 2 | Appropriate Test |
|-----------|-----------|----------------|
| Faculty (Science, Arts, Business) | Pass/Fail status | Chi-square test |
| Age (years) | GPA (0–4 scale) | Pearson or Spearman correlation |
| Gender (Male/Female) | Exam score (continuous) | Point-biserial correlation |
| Smoking (Yes/No) | Disease (Yes/No) | Chi-square or Fisher's exact |
| Income | Happiness (controlling for age) | Partial correlation |

: Example 4.1 — Test selection by variable type {.striped}

## R Example: Exploring Variable Relationships

```{r independence-intro}
# --- Use the built-in mtcars and titanic-style datasets ---
# Convert relevant mtcars variables to categorical
data(mtcars)
mtcars_cat <- mtcars |>
  mutate(
    transmission = factor(am, labels = c("Automatic", "Manual")),
    cylinders    = factor(cyl),
    efficiency   = factor(ifelse(mpg >= median(mpg), "High", "Low"),
                          levels = c("Low", "High"))
  )

# Quick overview of variable types
cat("=== Variable Type Overview ===\n")
cat("Categorical × Categorical: transmission vs. efficiency\n")
cat("Continuous × Continuous:   mpg vs. hp\n")
cat("Binary × Continuous:       transmission vs. mpg\n\n")

# Contingency table: transmission × efficiency
tab <- table(mtcars_cat$transmission, mtcars_cat$efficiency)
kable(addmargins(tab),
      caption = "Contingency Table: Transmission × Fuel Efficiency") |>
  kable_styling(bootstrap_options = c("striped","hover"), full_width = FALSE)
```

```{r independence-viz}
# --- Visualize potential associations ---
p1 <- ggplot(mtcars_cat,
             aes(x = transmission, fill = efficiency)) +
  geom_bar(position = "fill", color = "white") +
  scale_fill_manual(values = c("Low" = "tomato", "High" = "steelblue")) +
  scale_y_continuous(labels = scales::percent) +
  labs(title    = "Transmission vs. Fuel Efficiency",
       subtitle = "Stacked proportional bar — unequal proportions suggest association",
       x        = "Transmission Type",
       y        = "Proportion",
       fill     = "Efficiency") +
  theme_minimal(base_size = 12)

p2 <- ggplot(mtcars, aes(x = hp, y = mpg)) +
  geom_point(color = "steelblue", size = 2.5, alpha = 0.7) +
  geom_smooth(method = "lm", color = "tomato",
              se = TRUE, linewidth = 1) +
  labs(title    = "Horsepower vs. Fuel Efficiency",
       subtitle = "Scatterplot with linear fit — downward trend suggests negative correlation",
       x        = "Horsepower (hp)",
       y        = "Miles per Gallon (mpg)") +
  theme_minimal(base_size = 12)

p1 + p2
```

**Code explanation:**

- `addmargins(table())` adds row and column totals to the contingency table — essential for computing expected frequencies by hand.
- `geom_bar(position = "fill")` creates proportional stacked bars, making differences in conditional distributions visually obvious.
- Visual inspection always precedes formal testing — if proportions look identical across groups, a significant test result warrants careful scrutiny.

## Exercises

::: {.callout-tip}
## Exercise 4.1
For each of the following research questions, identify: (a) the variable types, (b) the appropriate test of independence, and (c) what a significant result would mean substantively.

1. Do male and female students differ in their choice of academic major?
2. Is there a relationship between study hours per week and final exam score?
3. Does coffee consumption (cups/day) relate to reported anxiety level (1–10 scale)?
4. Is being a first-generation student associated with dropout status (yes/no)?
:::

---

# Chi-Square Test of Independence

## Introduction

The **chi-square test of independence** is the standard procedure for testing whether two categorical variables are associated. It compares the **observed** frequencies in a contingency table to the **expected** frequencies that would arise if the variables were truly independent. It is among the most widely used tests in social science, medical research, and data science — applicable whenever we want to know if two categorical groupings are related.

## Theory

### The Test Statistic

Given a contingency table with $r$ rows and $c$ columns, the chi-square test statistic is:

$$\chi^2 = \sum_{i=1}^{r} \sum_{j=1}^{c} \frac{(O_{ij} - E_{ij})^2}{E_{ij}}$$

where $O_{ij}$ is the observed frequency and $E_{ij} = R_i C_j / N$ is the expected frequency in cell $(i,j)$.

Under $H_0$ (independence), $\chi^2 \sim \chi^2((r-1)(c-1))$ approximately.

**Hypotheses:**
$$H_0: \text{The two variables are independent}$$
$$H_1: \text{The two variables are associated}$$

### Assumptions

1. Observations are **independent** (each subject appears in exactly one cell).
2. The data are **counts** (not percentages or proportions).
3. **Expected cell frequencies** are sufficiently large:
   - At least 80% of cells have $E_{ij} \geq 5$
   - No cell has $E_{ij} < 1$

When these conditions are violated — typically with small samples or rare categories — use **Fisher's Exact Test** (Section 3).

### Yates' Continuity Correction

For 2×2 tables, a continuity correction is sometimes applied to account for the fact that we are approximating a discrete distribution with a continuous one:

$$\chi^2_{\text{Yates}} = \sum \frac{(|O_{ij} - E_{ij}| - 0.5)^2}{E_{ij}}$$

Yates' correction is conservative (reduces test power) and is debated in the literature. Many statisticians prefer Fisher's Exact Test for 2×2 tables regardless of sample size.

### Standardized Residuals

A significant chi-square test tells us *that* association exists but not *where* it comes from. **Standardized residuals** identify which cells contribute most:

$$r_{ij} = \frac{O_{ij} - E_{ij}}{\sqrt{E_{ij}}}$$

Cells with $|r_{ij}| > 2$ are typically considered notably discrepant from independence.

## Example: Chi-Square Test

**Example 4.2.** A researcher investigates whether exercise frequency is associated with self-reported health status among 200 adults.

|  | Poor Health | Good Health | Total |
|--|------------|------------|-------|
| Rarely exercises | 40 | 25 | 65 |
| Sometimes exercises | 30 | 45 | 75 |
| Regularly exercises | 15 | 45 | 60 |
| **Total** | **85** | **115** | **200** |

**Expected frequencies under $H_0$:**
$$E_{11} = \frac{65 \times 85}{200} = 27.6, \quad E_{12} = \frac{65 \times 115}{200} = 37.4$$
$$E_{21} = \frac{75 \times 85}{200} = 31.9, \quad E_{22} = \frac{75 \times 115}{200} = 43.1$$
$$E_{31} = \frac{60 \times 85}{200} = 25.5, \quad E_{32} = \frac{60 \times 115}{200} = 34.5$$

**Test statistic:**
$$\chi^2 = \frac{(40-27.6)^2}{27.6} + \frac{(25-37.4)^2}{37.4} + \cdots = 5.58 + 4.12 + 0.11 + 0.08 + 4.33 + 3.20 = 17.42$$

**df** $= (3-1)(2-1) = 2$, **p-value** $= P(\chi^2_2 > 17.42) < 0.001$

**Decision:** Reject $H_0$ — exercise frequency and health status are significantly associated ($\chi^2(2) = 17.42$, $p < 0.001$).

**Standardized residuals** reveal the pattern: "Rarely exercises / Poor Health" is notably higher than expected ($r = +2.36$), while "Regularly exercises / Poor Health" is notably lower ($r = -2.08$) — people who exercise regularly are less likely to report poor health.

## R Example: Chi-Square Test of Independence

```{r chisq}
# --- Build the exercise × health contingency table ---
exercise_health <- matrix(
  c(40, 25, 30, 45, 15, 45),
  nrow   = 3, byrow  = TRUE,
  dimnames = list(
    Exercise = c("Rarely", "Sometimes", "Regularly"),
    Health   = c("Poor", "Good")
  )
)

# Display observed table with margins
kable(addmargins(exercise_health),
      caption = "Observed Frequencies: Exercise × Health Status") |>
  kable_styling(bootstrap_options = c("striped","hover"),
                full_width = FALSE)
```

```{r chisq-test}
# --- Chi-square test ---
chi_result <- chisq.test(exercise_health, correct = FALSE)
print(chi_result)

# --- Expected frequencies ---
kable(round(chi_result$expected, 2),
      caption = "Expected Frequencies Under H₀ (Independence)") |>
  kable_styling(bootstrap_options = c("striped","hover"),
                full_width = FALSE)

# --- Standardized residuals ---
kable(round(chi_result$stdres, 3),
      caption = "Standardized Residuals (|r| > 2 = notable departure)") |>
  kable_styling(bootstrap_options = c("striped","hover"),
                full_width = FALSE) |>
  column_spec(2:3,
              color = ifelse(
                abs(chi_result$stdres) > 2, "tomato", "black"
              ))
```

```{r chisq-plot}
# --- Mosaic plot and association plot ---
# Convert to data frame for ggplot
eh_df <- as.table(exercise_health) |>
  as.data.frame(responseName = "Count")

# Proportional stacked bar
ggplot(eh_df, aes(x = Exercise, y = Count, fill = Health)) +
  geom_bar(stat = "identity", position = "fill", color = "white") +
  geom_hline(yintercept = 85/200, linetype = "dashed",
             color = "gray30", linewidth = 0.8) +
  annotate("text", x = 0.6, y = 85/200 + 0.03,
           label = "Overall poor health rate (42.5%)",
           color = "gray30", size = 3.5, hjust = 0) +
  scale_fill_manual(values = c("Poor" = "tomato", "Good" = "steelblue")) +
  scale_y_continuous(labels = scales::percent) +
  labs(title    = "Exercise Frequency vs. Health Status",
       subtitle = paste0("Chi-square test: χ²(2) = ",
                         round(chi_result$statistic, 2),
                         ", p < 0.001"),
       x        = "Exercise Frequency",
       y        = "Proportion",
       fill     = "Health Status") +
  theme_minimal(base_size = 13)
```

**Code explanation:**

- `chisq.test(table, correct = FALSE)` runs the chi-square test without Yates' correction. Set `correct = TRUE` for Yates' correction on 2×2 tables.
- The result object stores `$expected` (expected frequencies), `$stdres` (standardized residuals), and `$p.value`.
- The dashed horizontal line marks the overall "poor health" rate — bars above it indicate excess poor health relative to the population average, making deviations from independence visually clear.

**Try modifying:** Check `all(chi_result$expected >= 5)` to verify the minimum expected frequency assumption is met.

## Exercises

::: {.callout-tip}
## Exercise 4.2
Use the `HairEyeColor` dataset built into R (hair color × eye color × sex for 592 statistics students).

(a) Collapse across sex: `margin.table(HairEyeColor, c(1,2))` to get a Hair × Eye contingency table.
(b) Run a chi-square test. Are hair and eye color independent?
(c) Compute and display standardized residuals. Which hair-eye combinations are most over- or under-represented?
(d) Produce a proportional stacked bar chart and interpret the pattern.
:::

::: {.callout-tip}
## Exercise 4.3
A clinical study records treatment type (Drug A, Drug B, Placebo) and recovery outcome (Recovered, Not Recovered) for 150 patients.

(a) Create the contingency table in R with `matrix()`.
(b) Verify the minimum expected frequency assumption. If violated, what alternative test should you use?
(c) Run the chi-square test and report results in APA format: $\chi^2(df) = X.XX$, $p = .XXX$.
:::

---

# Fisher's Exact Test

## Introduction

The chi-square test relies on a large-sample approximation — when expected cell frequencies are small (below 5), the $\chi^2$ approximation becomes unreliable and p-values are inaccurate. **Fisher's Exact Test** solves this problem by computing the **exact** probability of observing the table, or one more extreme, under $H_0$ of independence. It was originally designed for 2×2 tables but extends to larger tables. In data science and biomedical research, small samples and rare events are common — Fisher's test is an essential tool.

## Theory

### The Hypergeometric Distribution

Fisher's test is based on the **hypergeometric distribution** — the exact probability of a specific 2×2 table with fixed marginal totals $(R_1, R_2, C_1, C_2)$:

$$P\left(\begin{bmatrix} a & b \\ c & d \end{bmatrix}\right) = \frac{\binom{R_1}{a}\binom{R_2}{c}}{\binom{N}{C_1}} = \frac{(R_1)!(R_2)!(C_1)!(C_2)!}{N! \cdot a! \cdot b! \cdot c! \cdot d!}$$

The p-value is the sum of probabilities for all tables as extreme as, or more extreme than, the observed table — while keeping marginal totals fixed.

### Two-Sided Fisher's Test

For two-sided tests, "more extreme" is defined as tables with probability **less than or equal to** the probability of the observed table. R's `fisher.test()` implements this by default.

### When to Use Fisher's Exact Test

| Situation | Recommendation |
|-----------|---------------|
| Any cell has $E_{ij} < 5$ | Use Fisher's |
| Total $N < 20$ | Use Fisher's |
| 2×2 table, any sample size | Fisher's is always valid (some prefer it over chi-square) |
| Large $r \times c$ table, large $N$ | Chi-square is adequate |

: Fisher's Exact Test usage guidelines {.striped}

## Example: Fisher's Exact Test

**Example 4.3.** A rare disease study examines whether a genetic marker is associated with disease status. Only 14 subjects are available.

|  | Disease | No Disease | Total |
|--|---------|-----------|-------|
| Marker Present | 5 | 2 | 7 |
| Marker Absent | 1 | 6 | 7 |
| **Total** | **6** | **8** | **14** |

**Expected frequencies:** $E_{11} = 6 \times 7/14 = 3.0$ — below 5, so chi-square is inappropriate.

**Fisher's exact p-value (two-sided):** $p = 0.103$

**Decision:** Fail to reject $H_0$ at $\alpha = 0.05$. Despite the apparent pattern, there is insufficient evidence to conclude the marker is associated with disease — likely due to the very small sample size.

**Interpretation note:** This example illustrates an important limitation. The study is severely **underpowered** — a larger study might well find significance. Fisher's test is honest about uncertainty; it does not manufacture significance from sparse data.

## R Example: Fisher's Exact Test

```{r fisher}
# --- Genetic marker × disease status ---
marker_disease <- matrix(
  c(5, 2, 1, 6),
  nrow     = 2, byrow    = TRUE,
  dimnames = list(
    Marker  = c("Present", "Absent"),
    Disease = c("Disease", "No Disease")
  )
)

kable(addmargins(marker_disease),
      caption = "Observed: Genetic Marker × Disease Status") |>
  kable_styling(bootstrap_options = c("striped","hover"),
                full_width = FALSE)

# --- Check expected frequencies ---
chi_check <- chisq.test(marker_disease, correct = FALSE)
cat("Expected frequencies:\n")
print(round(chi_check$expected, 2))
cat("\nMinimum expected frequency:",
    round(min(chi_check$expected), 2),
    "— chi-square assumption violated!\n\n")

# --- Fisher's Exact Test ---
fisher_result <- fisher.test(marker_disease, alternative = "two.sided")
print(fisher_result)
```

```{r fisher-compare}
# --- Compare Fisher's vs Chi-square on same data ---
comparison <- data.frame(
  Test      = c("Chi-Square (no correction)",
                "Chi-Square (Yates' correction)",
                "Fisher's Exact Test"),
  Statistic = c(round(chisq.test(marker_disease,
                                  correct=FALSE)$statistic, 3),
                round(chisq.test(marker_disease,
                                  correct=TRUE)$statistic, 3),
                "—"),
  p_value   = c(round(chisq.test(marker_disease,
                                   correct=FALSE)$p.value, 4),
                round(chisq.test(marker_disease,
                                   correct=TRUE)$p.value, 4),
                round(fisher_result$p.value, 4)),
  Appropriate = c("No (E < 5)", "No (E < 5)", "Yes ✓")
)

kable(comparison,
      caption   = "Test Comparison: Small Sample Scenario",
      col.names = c("Test", "Statistic", "p-value", "Appropriate?")) |>
  kable_styling(bootstrap_options = c("striped","hover"),
                full_width = FALSE) |>
  column_spec(4, bold = TRUE,
              color = c("tomato","tomato","darkgreen"))
```

```{r fisher-plot}
# --- Visualize with balloon plot ---
marker_df <- as.table(marker_disease) |>
  as.data.frame(responseName = "Count") 

ggplot(marker_df, aes(x = Disease, y = Marker)) +
  geom_point(aes(size = Count, color = Count),
             shape = 16, alpha = 0.8) +
  geom_text(aes(label = Count), color = "white",
            fontface = "bold", size = 5) +
  scale_size_continuous(range = c(8, 25)) +
  scale_color_gradient(low = "steelblue", high = "tomato") +
  labs(title    = "Balloon Plot: Marker × Disease",
       subtitle = paste0("Fisher's Exact Test: p = ",
                         round(fisher_result$p.value, 3)),
       x        = "Disease Status",
       y        = "Marker Status") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none")
```

**Code explanation:**

- `fisher.test(table, alternative)` computes the exact p-value. For 2×2 tables it also returns an **odds ratio** with a confidence interval — a useful effect measure.
- The comparison table illustrates how chi-square p-values are unreliable with small expected frequencies — they differ substantially from Fisher's exact result.
- The balloon plot is an intuitive alternative to a contingency table for small tables: bubble area represents cell count, making the pattern immediately visible.

## Exercises

::: {.callout-tip}
## Exercise 4.4
A new teaching intervention is tested in two small classrooms. Results:

|  | Passed | Failed |
|--|--------|--------|
| Intervention | 8 | 2 |
| Control | 4 | 6 |

(a) Check whether chi-square assumptions are met.
(b) Run Fisher's Exact Test. What is the p-value and odds ratio?
(c) Interpret the odds ratio in plain language.
(d) What would a larger study need to show to confirm this result?
:::

---

# Cramér's V and Effect Size for Categorical Association

## Introduction

A significant chi-square test only tells us that an association **exists** — it does not say how **strong** it is. With large samples, even trivially weak associations become statistically significant. **Effect size measures** for categorical variables answer the practical question: *How strongly are these two variables related?* The most versatile measure is **Cramér's V**, which standardizes chi-square to a 0–1 scale regardless of table dimensions.

## Theory

### Cramér's V

Cramér's V is derived from the chi-square statistic, normalized to fall between 0 and 1:

$$V = \sqrt{\frac{\chi^2 / N}{\min(r-1, c-1)}}$$

where $N$ is the total sample size, $r$ is the number of rows, and $c$ is the number of columns.

- $V = 0$: Perfect independence (no association)
- $V = 1$: Perfect association

**Interpretation benchmarks** (Cohen, 1988), adjusted for table size:

| Table Size | Small | Medium | Large |
|-----------|-------|--------|-------|
| 2×2 | 0.10 | 0.30 | 0.50 |
| 3×3 | 0.07 | 0.21 | 0.35 |
| 4×4 | 0.06 | 0.17 | 0.29 |

: Cramér's V interpretation by table dimensions {.striped}

### The Phi Coefficient

For 2×2 tables specifically, the **phi coefficient** $\phi$ is a simpler equivalent:

$$\phi = \sqrt{\frac{\chi^2}{N}} = \frac{ad - bc}{\sqrt{(a+b)(c+d)(a+c)(b+d)}}$$

$\phi$ ranges from $-1$ to $+1$ for 2×2 tables (with sign indicating direction) and equals Cramér's V when $\min(r-1, c-1) = 1$.

### Goodman and Kruskal's Lambda

An alternative measure of association is **Lambda** ($\lambda$), which measures the **proportional reduction in prediction error** — how much better we can predict one variable knowing the other:

$$\lambda = \frac{\sum_j \max_i f_{ij} - \max_i R_i}{N - \max_i R_i}$$

$\lambda = 0$ means knowing $X$ does not help predict $Y$; $\lambda = 1$ means perfect prediction.

## Example: Cramér's V

**Example 4.4.** Returning to the exercise × health study (Example 4.2): $\chi^2 = 17.42$, $N = 200$, table is 3×2.

$$V = \sqrt{\frac{17.42/200}{\min(3-1, 2-1)}} = \sqrt{\frac{0.0871}{1}} = \sqrt{0.0871} \approx 0.295$$

For a table with $\min(r-1, c-1) = 1$, benchmarks are 0.10/0.30/0.50 (small/medium/large). $V = 0.295$ is close to a **medium effect** — the association between exercise and health is statistically significant and of moderate practical importance.

## R Example: Cramér's V and Association Measures

```{r cramers-v}
# --- Cramér's V for exercise × health ---
exercise_health <- matrix(
  c(40, 25, 30, 45, 15, 45),
  nrow = 3, byrow = TRUE,
  dimnames = list(
    Exercise = c("Rarely", "Sometimes", "Regularly"),
    Health   = c("Poor", "Good")
  )
)

chi_result  <- chisq.test(exercise_health, correct = FALSE)
N           <- sum(exercise_health)
min_dim     <- min(nrow(exercise_health)-1, ncol(exercise_health)-1)
cramers_v   <- sqrt(chi_result$statistic / (N * min_dim))

cat("Chi-square statistic:", round(chi_result$statistic, 3), "\n")
cat("N:", N, "\n")
cat("min(r-1, c-1):", min_dim, "\n")
cat("Cramér's V:", round(cramers_v, 4), "\n")
cat("Interpretation: Medium effect (benchmark ~0.30 for this table size)\n\n")

# --- Using vcd package for association measures ---
assoc_stats <- vcd::assocstats(exercise_health)
print(assoc_stats)
```

```{r cramers-heatmap}
# --- Visualize standardized residuals as heatmap ---
stdres <- chi_result$stdres
stdres_df <- as.data.frame(as.table(stdres)) |>
  rename(Exercise = Exercise, Health = Health, Residual = Freq)

ggplot(stdres_df, aes(x = Health, y = Exercise, fill = Residual)) +
  geom_tile(color = "white", linewidth = 1.5) +
  geom_text(aes(label = round(Residual, 2)),
            size = 5, fontface = "bold",
            color = ifelse(abs(stdres_df$Residual) > 2,
                           "white", "black")) +
  scale_fill_gradient2(low  = "steelblue", mid = "white",
                       high = "tomato",    midpoint = 0,
                       limits = c(-3, 3)) +
  labs(title    = "Standardized Residuals Heatmap",
       subtitle = paste0("Cramér's V = ", round(cramers_v, 3),
                         " (medium effect) | Red = more than expected, Blue = less"),
       x        = "Health Status",
       y        = "Exercise Frequency",
       fill     = "Std. Residual") +
  theme_minimal(base_size = 13)
```

**Code explanation:**

- Cramér's V is computed manually here to show the formula clearly; `vcd::assocstats()` computes it (and phi, contingency coefficient) in one call.
- The heatmap of standardized residuals is more informative than a standard mosaic plot for larger tables — color immediately draws the eye to cells with the strongest departures from independence.
- `scale_fill_gradient2()` creates a diverging color scale centered at zero, perfect for signed residuals.

## Exercises

::: {.callout-tip}
## Exercise 4.5
Using the `HairEyeColor` chi-square result from Exercise 4.2:

(a) Compute Cramér's V manually from the $\chi^2$ statistic.
(b) Use `vcd::assocstats()` to verify.
(c) Interpret the effect size using the appropriate benchmark for the table dimensions.
(d) Produce a standardized residuals heatmap and describe which hair-eye combinations drive the association.
:::

---

# Correlation Tests

## Introduction

When both variables are continuous (or at least ordinal with many levels), **correlation** quantifies the strength and direction of their linear relationship. But a sample correlation $r$ could arise by chance even when the true population correlation $\rho = 0$. A **correlation test** determines whether the observed correlation is statistically distinguishable from zero. In this section we cover the two most important correlation measures: **Pearson's $r$** for linear relationships between normally distributed variables, and **Spearman's $\rho$** for monotone relationships with no normality requirement.

## Theory

### Pearson Correlation Test

**Hypotheses:** $H_0: \rho = 0$ vs. $H_1: \rho \neq 0$ (or one-sided)

Given sample correlation $r$ from $n$ pairs, the test statistic is:

$$t = \frac{r\sqrt{n-2}}{\sqrt{1-r^2}} \sim t(n-2) \text{ under } H_0$$

**Assumptions:**

1. Both variables are continuous.
2. The relationship is **linear**.
3. Both variables are approximately **normally distributed** (bivariate normality).
4. Observations are independent.
5. No extreme outliers (Pearson's $r$ is sensitive to outliers).

**Confidence interval for $\rho$:** Use Fisher's z-transformation:
$$z = \frac{1}{2}\ln\!\left(\frac{1+r}{1-r}\right), \quad \text{SE}(z) = \frac{1}{\sqrt{n-3}}$$

Then transform back to the correlation scale.

### Spearman Rank Correlation Test

**Spearman's $\rho$** replaces raw values with **ranks** and computes Pearson's $r$ on the ranks. This makes it robust to:

- Non-normality
- Outliers
- Monotone but non-linear relationships
- Ordinal variables

$$\rho_s = 1 - \frac{6\sum d_i^2}{n(n^2-1)}$$

where $d_i$ is the difference in ranks for pair $i$. For large samples, a t-approximation is used for hypothesis testing.

### Interpreting Correlation

| $|r|$ or $|\rho|$ | Strength |
|-------------------|----------|
| 0.00 – 0.19 | Negligible |
| 0.20 – 0.39 | Weak |
| 0.40 – 0.59 | Moderate |
| 0.60 – 0.79 | Strong |
| 0.80 – 1.00 | Very strong |

: Correlation strength benchmarks {.striped}

::: {.callout-warning}
## Critical Reminder: Correlation ≠ Causation
A significant correlation between $X$ and $Y$ does not imply that $X$ causes $Y$, nor that $Y$ causes $X$. Both could be driven by a third variable $Z$ (confounding), or the association could be spurious. Always consider alternative explanations and visualize the scatterplot alongside the correlation coefficient.
:::

## Example: Pearson and Spearman Correlation Tests

**Example 4.5.** A researcher examines the relationship between hours of sleep per night ($X$) and cognitive performance score ($Y$) in 40 participants.

Results: $r = 0.61$, $t_{38} = 4.73$, $p < 0.001$, 95% CI: $(0.37, 0.78)$.

**Interpretation:** There is a statistically significant, strong positive linear correlation between sleep duration and cognitive performance ($r = 0.61$, $p < 0.001$). The 95% CI $(0.37, 0.78)$ excludes zero, confirming the association. However, this is an observational study — we cannot conclude that more sleep **causes** better performance.

## R Example: Pearson and Spearman Correlation Tests

```{r correlation}
# --- Use mtcars: test correlation between hp and mpg ---
data(mtcars)

# Pearson correlation test
pearson_result <- cor.test(mtcars$hp, mtcars$mpg,
                            method      = "pearson",
                            alternative = "two.sided")

# Spearman correlation test
spearman_result <- cor.test(mtcars$hp, mtcars$mpg,
                             method      = "spearman",
                             alternative = "two.sided",
                             exact       = FALSE)

# Summary table
cor_summary <- data.frame(
  Method    = c("Pearson", "Spearman"),
  r_or_rho  = round(c(pearson_result$estimate,
                       spearman_result$estimate), 4),
  Statistic = round(c(pearson_result$statistic,
                       spearman_result$statistic), 3),
  df        = c(pearson_result$parameter, "—"),
  p_value   = round(c(pearson_result$p.value,
                       spearman_result$p.value), 6),
  CI_lower  = c(round(pearson_result$conf.int[1], 3), "—"),
  CI_upper  = c(round(pearson_result$conf.int[2], 3), "—")
)

kable(cor_summary,
      caption   = "Correlation Test Results: Horsepower vs. MPG",
      col.names = c("Method", "r / ρ", "t / S",
                    "df", "p-value", "95% CI Lower", "95% CI Upper")) |>
  kable_styling(bootstrap_options = c("striped","hover"),
                full_width = FALSE)
```

```{r correlation-plot}
# --- Scatterplot with correlation annotation ---
r_val <- round(pearson_result$estimate, 3)
p_val <- round(pearson_result$p.value, 4)

p1 <- ggplot(mtcars, aes(x = hp, y = mpg)) +
  geom_point(color = "steelblue", size = 2.5, alpha = 0.8) +
  geom_smooth(method = "lm", color = "tomato",
              se = TRUE, linewidth = 1.1) +
  annotate("text", x = 280, y = 30,
           label = paste0("r = ", r_val,
                          "\np = ", p_val),
           color = "tomato", fontface = "bold",
           size = 4, hjust = 1) +
  labs(title    = "Horsepower vs. Fuel Efficiency",
       subtitle = "Pearson correlation with 95% confidence band",
       x        = "Horsepower (hp)",
       y        = "Miles per Gallon (mpg)") +
  theme_minimal(base_size = 12)

# --- Full correlation matrix for mtcars ---
cont_vars <- mtcars[, c("mpg","cyl","disp","hp","wt","qsec")]
cor_mat   <- cor(cont_vars, method = "pearson")

p2 <- ggcorrplot(cor_mat,
                 method    = "circle",
                 type      = "upper",
                 lab       = TRUE,
                 lab_size  = 3.5,
                 colors    = c("tomato","white","steelblue"),
                 title     = "Correlation Matrix — mtcars",
                 ggtheme   = theme_minimal(base_size = 11))

p1 + p2
```

**Code explanation:**

- `cor.test(x, y, method, alternative)` tests both Pearson (`"pearson"`) and Spearman (`"spearman"`) correlations. Pearson returns a CI; Spearman does not (due to the rank-based nature).
- `exact = FALSE` in Spearman uses a normal approximation — necessary when ties are present in the data.
- `ggcorrplot()` produces a publication-ready correlation matrix. `type = "upper"` shows only the upper triangle to avoid redundancy.

## Exercises

::: {.callout-tip}
## Exercise 4.6
Using the `iris` dataset, test the correlation between `Sepal.Length` and `Petal.Length`.

(a) Produce a scatterplot. Does the relationship appear linear?
(b) Check normality of both variables (Shapiro-Wilk). Which correlation method is more appropriate?
(c) Run both Pearson and Spearman tests. Compare and explain any differences.
(d) Compute the full 4×4 correlation matrix for all numeric variables and visualize with `ggcorrplot()`.
:::

::: {.callout-tip}
## Exercise 4.7
The `anscombe` dataset contains four pairs of variables with nearly identical correlations but very different scatterplot shapes.

(a) Compute Pearson's $r$ for all four pairs (x1/y1, x2/y2, x3/y3, x4/y4).
(b) Plot all four scatterplots in a 2×2 grid.
(c) For which pairs is Pearson's $r$ a misleading summary? What would be more appropriate?
:::

---

# Point-Biserial and Phi Coefficients

## Introduction

Not all association problems involve two continuous or two categorical variables. In data science, binary variables are ubiquitous — pass/fail, churn/retain, click/no-click, disease/healthy. When one or both variables in a pair are binary, specialized correlation measures maintain interpretability while remaining mathematically consistent with the general correlation framework. The **point-biserial correlation** handles the binary-continuous case; the **phi coefficient** handles the binary-binary case.

## Theory

### Point-Biserial Correlation ($r_{pb}$)

The **point-biserial correlation** measures the association between a **naturally binary variable** (coded 0/1) and a **continuous variable**. It is mathematically equivalent to Pearson's $r$ computed between the 0/1 binary variable and the continuous variable:

$$r_{pb} = \frac{\bar{Y}_1 - \bar{Y}_0}{s_Y}\sqrt{\frac{n_1 n_0}{n^2}}$$

where $\bar{Y}_1$ and $\bar{Y}_0$ are the means of the continuous variable in the two groups, $s_Y$ is the overall standard deviation, and $n_1$, $n_0$ are the group sizes.

**Significance testing:** Because $r_{pb}$ is algebraically equivalent to Pearson's $r$, the same t-test applies:

$$t = \frac{r_{pb}\sqrt{n-2}}{\sqrt{1-r_{pb}^2}} \sim t(n-2)$$

This is equivalent to running an **independent samples t-test** comparing the two groups — the same information, expressed differently.

**Assumptions:** The continuous variable should be approximately normally distributed within each group (same as the independent samples t-test).

### Phi Coefficient ($\phi$)

The **phi coefficient** measures association between **two binary variables**. It is Pearson's $r$ applied to two 0/1 variables:

$$\phi = \frac{ad - bc}{\sqrt{(a+b)(c+d)(a+c)(b+d)}}$$

where $a, b, c, d$ are the four cells of a 2×2 contingency table. $\phi$ ranges from $-1$ to $+1$ and equals Cramér's V for 2×2 tables.

**Interpretation benchmarks** (same as Pearson's $r$):

- $|\phi| < 0.10$: Negligible
- $0.10 \leq |\phi| < 0.30$: Weak
- $0.30 \leq |\phi| < 0.50$: Moderate
- $|\phi| \geq 0.50$: Strong

### Biserial vs. Point-Biserial

A subtle distinction: the **point-biserial** is used when the binary variable is *truly* dichotomous (e.g., sex, treatment assignment). The **biserial correlation** is used when the binary variable is an artificially dichotomized continuous variable (e.g., pass/fail from a continuous score). The biserial correlation is larger in magnitude and corrects for the information lost by dichotomization.

## Example: Point-Biserial Correlation

**Example 4.6.** A study records whether students passed a preparatory course (0 = No, 1 = Yes) and their final exam score. Results: $r_{pb} = 0.48$, $t_{58} = 4.19$, $p < 0.001$.

**Interpretation:** Students who passed the preparatory course scored significantly higher on the final exam ($r_{pb} = 0.48$, $p < 0.001$). This is a moderate-to-strong effect, indicating the preparatory course is associated with better exam performance. Note that this is equivalent to a two-sample t-test comparing exam scores between the two groups.

## R Example: Point-Biserial and Phi Coefficients

```{r point-biserial}
# --- Point-biserial: transmission (binary) vs. mpg (continuous) ---
# am: 0 = automatic, 1 = manual
data(mtcars)

# Method 1: cor.test (Pearson on binary × continuous)
pb_result <- cor.test(mtcars$am, mtcars$mpg,
                       method = "pearson")

cat("=== Point-Biserial Correlation ===\n")
cat("r_pb:", round(pb_result$estimate, 4), "\n")
cat("t:   ", round(pb_result$statistic, 3), "\n")
cat("df:  ", pb_result$parameter, "\n")
cat("p:   ", round(pb_result$p.value, 4), "\n")
cat("95% CI: [", round(pb_result$conf.int[1],3),
    ",", round(pb_result$conf.int[2],3), "]\n\n")

# Equivalence to t-test
t_equiv <- t.test(mpg ~ am, data = mtcars, var.equal = FALSE)
cat("Equivalent Welch's t-test p-value:",
    round(t_equiv$p.value, 4), "(same conclusion)\n")
```

```{r phi-coeff}
# --- Phi coefficient: am (auto/manual) × efficiency (high/low) ---
mtcars_bin <- mtcars |>
  mutate(
    transmission = factor(am, labels = c("Auto","Manual")),
    efficiency   = factor(ifelse(mpg >= median(mpg),
                                  "High","Low"),
                           levels = c("Low","High"))
  )

tab_2x2 <- table(mtcars_bin$transmission,
                  mtcars_bin$efficiency)

# Phi coefficient
chi_2x2 <- chisq.test(tab_2x2, correct = FALSE)
phi_val  <- sqrt(chi_2x2$statistic / sum(tab_2x2))

cat("=== Phi Coefficient ===\n")
kable(addmargins(tab_2x2),
      caption = "Transmission × Efficiency (2×2 Table)") |>
  kable_styling(bootstrap_options = c("striped","hover"),
                full_width = FALSE)

cat("Phi coefficient:", round(phi_val, 4), "\n")
cat("Chi-square p-value:", round(chi_2x2$p.value, 4), "\n")
cat("Interpretation: Moderate-to-strong association\n")
```

```{r pb-plot}
# --- Visualize point-biserial ---
p1 <- ggplot(mtcars_bin, aes(x = transmission,
                              y = mpg, fill = transmission)) +
  geom_violin(alpha = 0.5, trim = FALSE) +
  geom_boxplot(width = 0.12, fill = "white") +
  scale_fill_manual(values = c("Auto" = "steelblue",
                                "Manual" = "tomato")) +
  annotate("text", x = 1.5, y = 35,
           label = paste0("r_pb = ",
                           round(pb_result$estimate, 3),
                           "\np = ", round(pb_result$p.value, 4)),
           size = 4, fontface = "bold", color = "gray20") +
  labs(title = "A. Point-Biserial: Transmission vs. MPG",
       x = "Transmission", y = "MPG") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none")

# --- Visualize phi ---
p2 <- as.data.frame(tab_2x2) |>
  rename(Transmission = Var1, Efficiency = Var2) |>
  ggplot(aes(x = Efficiency, y = Transmission,
             fill = Freq)) +
  geom_tile(color = "white", linewidth = 1.5) +
  geom_text(aes(label = Freq), size = 6, fontface = "bold") +
  scale_fill_gradient(low = "white", high = "steelblue") +
  annotate("text", x = 1.5, y = 2.45,
           label = paste0("φ = ", round(phi_val, 3)),
           size  = 4.5, fontface = "bold") +
  labs(title = "B. Phi Coefficient: Transmission × Efficiency",
       x = "Efficiency", y = "Transmission") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none")

p1 + p2
```

**Code explanation:**

- `cor.test(binary, continuous, method = "pearson")` correctly computes the point-biserial correlation — no special function is needed since it is mathematically Pearson's $r$.
- Phi is computed as $\sqrt{\chi^2/N}$ — identical to Cramér's V for 2×2 tables.
- The equivalence between point-biserial and the independent samples t-test is shown explicitly — both tests use the same information and yield the same p-value (up to the choice of equal/unequal variances).

## Exercises

::: {.callout-tip}
## Exercise 4.8
Using the `mtcars` dataset:

(a) Compute the point-biserial correlation between `vs` (engine shape: 0 = V-shaped, 1 = straight) and `qsec` (quarter-mile time).
(b) Verify this equals the Pearson correlation between the two variables.
(c) Is there a significant association? Report in full.
(d) Compute the phi coefficient between `vs` and `am` (both binary). What is the strength of this association?
:::

---

# Partial Correlation

## Introduction

Simple bivariate correlation measures the relationship between two variables, but in real datasets variables are rarely isolated — many are influenced by common underlying factors. **Partial correlation** measures the association between two variables $X$ and $Y$ **after removing the influence of one or more control variables $Z$**. It answers: *Is the relationship between $X$ and $Y$ real, or is it merely a reflection of both being related to $Z$?* This is a crucial step in distinguishing genuine associations from spurious ones, and a natural bridge toward multiple regression (Chapter 10).

## Theory

### First-Order Partial Correlation

The **first-order partial correlation** between $X$ and $Y$ controlling for $Z$ is:

$$r_{XY \cdot Z} = \frac{r_{XY} - r_{XZ} \cdot r_{YZ}}{\sqrt{(1-r_{XZ}^2)(1-r_{YZ}^2)}}$$

where $r_{XY}$, $r_{XZ}$, $r_{YZ}$ are the zero-order (simple) Pearson correlations.

**Interpretation:** $r_{XY \cdot Z}$ is the correlation between the **residuals** of regressing $X$ on $Z$ and the **residuals** of regressing $Y$ on $Z$ — it is the "pure" relationship between $X$ and $Y$ once the linear effect of $Z$ is removed from both.

### Significance Testing

The partial correlation is tested using a t-distribution with $n - 2 - k$ degrees of freedom, where $k$ is the number of control variables:

$$t = \frac{r_{XY \cdot Z}\sqrt{n-2-k}}{\sqrt{1-r_{XY \cdot Z}^2}} \sim t(n-2-k)$$

### Partial vs. Semi-Partial Correlation

A related concept is the **semi-partial (part) correlation** $r_{X(Y \cdot Z)}$: the correlation between $X$ and the part of $Y$ that is **unique** (not shared with $Z$). It measures the unique contribution of $Y$ to $X$, which is central to multiple regression (it is the square root of the $\Delta R^2$ when $Y$ is added to a regression already containing $Z$).

### Spurious Correlation

A **spurious correlation** occurs when $X$ and $Y$ are correlated only because both are caused by a third variable $Z$. After partialling out $Z$, the partial correlation drops to (near) zero. Classic example: ice cream sales and drowning rates are positively correlated — both are caused by hot weather (the confounder). The partial correlation controlling for temperature would be near zero.

## Example: Partial Correlation

**Example 4.7.** In the `mtcars` dataset, `mpg` and `hp` are strongly negatively correlated ($r = -0.78$). But both are also related to car weight (`wt`): heavier cars tend to have more horsepower and worse fuel economy. Is the `mpg`-`hp` correlation "real" or partly a reflection of their shared relationship with `wt`?

- $r_{\text{mpg,hp}} = -0.78$
- $r_{\text{mpg,wt}} = -0.87$
- $r_{\text{hp,wt}} = +0.66$

**Partial correlation controlling for `wt`:**

$$r_{\text{mpg,hp} \cdot \text{wt}} = \frac{-0.78 - (-0.87)(0.66)}{\sqrt{(1-0.87^2)(1-0.66^2)}} = \frac{-0.78 + 0.574}{\sqrt{0.243 \times 0.564}} = \frac{-0.206}{0.370} \approx -0.557$$

The partial correlation ($-0.56$) is notably weaker than the zero-order correlation ($-0.78$), suggesting that part of the `mpg`-`hp` relationship is mediated through vehicle weight. However, a meaningful negative relationship remains even after controlling for weight.

## R Example: Partial Correlation

```{r partial-corr}
# --- Partial correlations in mtcars ---
data(mtcars)

# Zero-order correlations (for context)
cat("=== Zero-Order Correlations ===\n")
cor_mat <- cor(mtcars[, c("mpg","hp","wt","disp")])
round(cor_mat, 3)
```

```{r partial-corr2}
# --- Partial correlation: mpg ~ hp controlling for wt ---
library(ppcor)

# First-order partial correlation (controlling for one variable)
partial_result <- pcor.test(mtcars$mpg, mtcars$hp, mtcars$wt)

cat("=== Partial Correlation: mpg & hp | wt ===\n")
cat("Zero-order r(mpg, hp):     ",
    round(cor(mtcars$mpg, mtcars$hp), 4), "\n")
cat("Partial r(mpg, hp | wt):   ",
    round(partial_result$estimate, 4), "\n")
cat("t-statistic:               ",
    round(partial_result$statistic, 3), "\n")
cat("df:                        ", partial_result$n - 2 - 1, "\n")
cat("p-value:                   ",
    round(partial_result$p.value, 4), "\n\n")

# Full partial correlation matrix (controlling for wt)
pcor_matrix <- pcor(mtcars[, c("mpg","hp","disp","qsec")],
                    method = "pearson")

cat("=== Partial Correlation Matrix (all pairs, no controls) ===\n")
round(pcor_matrix$estimate, 3)
```

```{r partial-plot}
# --- Compare zero-order vs partial correlation visually ---
# Compute residuals of mpg and hp after removing wt's influence
resid_mpg <- residuals(lm(mpg ~ wt, data = mtcars))
resid_hp  <- residuals(lm(hp  ~ wt, data = mtcars))

p1 <- ggplot(mtcars, aes(x = hp, y = mpg)) +
  geom_point(color = "steelblue", size = 2.5, alpha = 0.8) +
  geom_smooth(method = "lm", color = "tomato",
              se = FALSE, linewidth = 1) +
  annotate("text", x = 280, y = 31,
           label = paste0("r = ",
                           round(cor(mtcars$hp, mtcars$mpg), 3)),
           color = "tomato", fontface = "bold", size = 4) +
  labs(title    = "A. Zero-Order Correlation",
       subtitle = "mpg vs. hp (no control)",
       x = "Horsepower (hp)", y = "MPG") +
  theme_minimal(base_size = 12)

p2 <- ggplot(data.frame(res_hp  = resid_hp,
                         res_mpg = resid_mpg),
             aes(x = res_hp, y = res_mpg)) +
  geom_point(color = "seagreen", size = 2.5, alpha = 0.8) +
  geom_smooth(method = "lm", color = "tomato",
              se = FALSE, linewidth = 1) +
  annotate("text", x = 100, y = 4,
           label = paste0("r = ",
                           round(cor(resid_hp, resid_mpg), 3)),
           color = "tomato", fontface = "bold", size = 4) +
  labs(title    = "B. Partial Correlation",
       subtitle = "mpg vs. hp (after removing wt influence)",
       x = "hp Residual (after wt)",
       y = "mpg Residual (after wt)") +
  theme_minimal(base_size = 12)

p1 + p2
```

**Code explanation:**

- `pcor.test(x, y, z)` from `ppcor` computes the first-order partial correlation between `x` and `y` controlling for `z`, with a significance test.
- `pcor(data_frame)` computes the full partial correlation matrix (each pair controlled for all others) — analogous to `cor()` for zero-order correlations.
- Panels A and B illustrate the concept geometrically: Panel B shows the scatterplot of **residuals** — the pure association between `mpg` and `hp` after both have been "de-weighted" by `wt`.

**Try modifying:** Control for both `wt` and `disp` simultaneously using `pcor.test(mpg, hp, cbind(wt, disp))`. Does the partial correlation weaken further?

## Exercises

::: {.callout-tip}
## Exercise 4.9
Using the `iris` dataset, `Sepal.Length` and `Petal.Length` are strongly correlated ($r \approx 0.87$). Species is a potential confounder.

(a) Compute the zero-order Pearson correlation between `Sepal.Length` and `Petal.Length`.
(b) Encode `Species` as a numeric variable (1, 2, 3) and compute the partial correlation controlling for it using `pcor.test()`.
(c) Does the partial correlation differ substantially from the zero-order correlation? What does this suggest?
(d) Visualize both zero-order and partial relationships as scatterplots (color points by species in panel A).
:::

::: {.callout-tip}
## Exercise 4.10 (Challenge)
Using `mtcars`, construct a full analysis comparing zero-order and partial correlations for the variable pair (`wt`, `mpg`), controlling for `cyl` (number of cylinders).

(a) Compute zero-order $r$(wt, mpg) and partial $r$(wt, mpg | cyl).
(b) Do the correlations differ? What does this tell us about the role of engine size?
(c) Produce a 2-panel residual scatterplot similar to the chapter example.
(d) Write a 150-word research summary reporting both correlations in APA format.
:::

---

# Chapter Lab Activity: Exploring Independence with the `titanic` and `mtcars` Datasets

## Objectives

In this lab you will apply the full spectrum of independence tests to real data — choosing the correct test based on variable types, checking assumptions, reporting effect sizes, and visualizing associations. You will work with the built-in `Titanic` dataset (categorical variables) and `mtcars` (mixed variable types).

## The Datasets

```{r lab-intro}
# --- Titanic dataset (built-in) ---
titanic_df <- as.data.frame(Titanic)
titanic_expanded <- titanic_df[rep(seq_len(nrow(titanic_df)),
                                    titanic_df$Freq), -5]
rownames(titanic_expanded) <- NULL

kable(head(titanic_expanded, 10),
      caption = "Titanic Dataset (first 10 rows)") |>
  kable_styling(bootstrap_options = c("striped","hover"),
                font_size = 11)

cat("Total passengers:", nrow(titanic_expanded), "\n")
cat("Variables:", names(titanic_expanded), "\n")
```

## Lab Task 1: Survival × Class — Chi-Square Test

```{r lab-task1}
# Contingency table: Class × Survived
class_surv <- table(titanic_expanded$Class,
                     titanic_expanded$Survived)

kable(addmargins(class_surv),
      caption = "Passenger Class × Survival Status") |>
  kable_styling(bootstrap_options = c("striped","hover"),
                full_width = FALSE)

# Chi-square test
chi_titanic <- chisq.test(class_surv, correct = FALSE)
cat("\nChi-Square Test Results:\n")
cat("χ²(", chi_titanic$parameter, ") =",
    round(chi_titanic$statistic, 2), "\n")
cat("p-value:", round(chi_titanic$p.value, 6), "\n")
cat("All expected ≥ 5:", all(chi_titanic$expected >= 5), "\n")

# Cramér's V
N_t     <- sum(class_surv)
min_dim <- min(nrow(class_surv)-1, ncol(class_surv)-1)
v_t     <- sqrt(chi_titanic$statistic / (N_t * min_dim))
cat("Cramér's V:", round(v_t, 4), "\n")
```

## Lab Task 2: Survival × Sex — Fisher's Exact Test

```{r lab-task2}
# Sex × Survived (2×2 table)
sex_surv <- table(titanic_expanded$Sex,
                   titanic_expanded$Survived)

kable(addmargins(sex_surv),
      caption = "Sex × Survival Status") |>
  kable_styling(bootstrap_options = c("striped","hover"),
                full_width = FALSE)

# Both chi-square and Fisher's for comparison
chi_sex    <- chisq.test(sex_surv, correct = FALSE)
fisher_sex <- fisher.test(sex_surv)

phi_sex <- sqrt(chi_sex$statistic / sum(sex_surv))

cat("Chi-Square: χ²(1) =", round(chi_sex$statistic, 2),
    ", p =", round(chi_sex$p.value, 6), "\n")
cat("Fisher's Exact: p =", round(fisher_sex$p.value, 6), "\n")
cat("Odds Ratio:", round(fisher_sex$estimate, 3),
    "  95% CI: [",
    round(fisher_sex$conf.int[1], 3), ",",
    round(fisher_sex$conf.int[2], 3), "]\n")
cat("Phi coefficient:", round(phi_sex, 4), "\n")
```

## Lab Task 3: Correlation Analysis — mtcars

```{r lab-task3}
# Full correlation matrix with significance tests
data(mtcars)
cont_vars <- mtcars[, c("mpg","hp","wt","disp","qsec")]

# Correlation matrix
cor_mat  <- cor(cont_vars)
#p_mat    <- cor_pmat(cont_vars)   # p-value matrix from ggcorrplot
ggcorrplot(cor_mat,
           method    = "circle",
           type      = "upper",
           lab       = TRUE,
           sig.level = 0.05,
           lab_size  = 3.5,
           insig     = "blank",
           colors    = c("tomato","white","steelblue"),
           title     = "Correlation Matrix: mtcars\n(blanked = not significant at α=0.05)",
           ggtheme   = theme_minimal(base_size = 11))

```

## Lab Task 4: Partial Correlation — Controlling for Weight

```{r lab-task4}
# Compare zero-order vs partial correlations
pairs_to_test <- list(
  c("mpg", "hp"),
  c("mpg", "disp"),
  c("hp",  "disp")
)

partial_summary <- map_dfr(pairs_to_test, function(pair) {
  x <- mtcars[[pair[1]]]
  y <- mtcars[[pair[2]]]
  z <- mtcars$wt

  r0  <- cor(x, y)
  res <- pcor.test(x, y, z)

  data.frame(
    Pair           = paste(pair[1], "~", pair[2]),
    Zero_order_r   = round(r0, 4),
    Partial_r_wt   = round(res$estimate, 4),
    p_value        = round(res$p.value, 4),
    Change         = round(res$estimate - r0, 4)
  )
})

kable(partial_summary,
      caption   = "Zero-Order vs. Partial Correlations (controlling for wt)",
      col.names = c("Variable Pair", "Zero-Order r",
                    "Partial r | wt", "p-value",
                    "Change in r")) |>
  kable_styling(bootstrap_options = c("striped","hover")) |>
  column_spec(5, bold = TRUE,
              color = ifelse(abs(partial_summary$Change) > 0.1,
                             "tomato", "gray40"))
```

## Lab Task 5: Comprehensive Visualization

```{r lab-task5, fig.height=8}
# Panel A: Survival rates by class (Titanic)
p1 <- as.data.frame(class_surv) |>
  rename(Class = Var1, Survived = Var2) |>
  group_by(Class) |>
  mutate(pct = Freq / sum(Freq)) |>
  ggplot(aes(x = Class, y = pct, fill = Survived)) +
  geom_col(color = "white") +
  scale_fill_manual(values = c("No" = "tomato",
                                "Yes" = "steelblue")) +
  scale_y_continuous(labels = scales::percent) +
  labs(title = "A. Survival Rate by Passenger Class",
       subtitle = paste0("Cramér's V = ", round(v_t, 3)),
       x = "Class", y = "Proportion") +
  theme_minimal(base_size = 11)

# Panel B: Survival by sex
p2 <- as.data.frame(sex_surv) |>
  rename(Sex = Var1, Survived = Var2) |>
  group_by(Sex) |>
  mutate(pct = Freq / sum(Freq)) |>
  ggplot(aes(x = Sex, y = pct, fill = Survived)) +
  geom_col(color = "white") +
  scale_fill_manual(values = c("No" = "tomato",
                                "Yes" = "steelblue")) +
  scale_y_continuous(labels = scales::percent) +
  labs(title = "B. Survival Rate by Sex",
       subtitle = paste0("φ = ", round(phi_sex, 3),
                         " | OR = ",
                         round(fisher_sex$estimate, 2)),
       x = "Sex", y = "Proportion") +
  theme_minimal(base_size = 11)

# Panel C: Standardized residuals heatmap (Class × Survived)
stdres_df <- as.data.frame(chi_titanic$stdres) |>
  rename(Class = Var1, Survived = Var2, Residual = Freq)

p3 <- ggplot(stdres_df,
             aes(x = Survived, y = Class, fill = Residual)) +
  geom_tile(color = "white", linewidth = 1.2) +
  geom_text(aes(label = round(Residual, 2)),
            fontface = "bold", size = 4.5,
            color = ifelse(abs(stdres_df$Residual) > 2,
                           "white", "black")) +
  scale_fill_gradient2(low  = "steelblue", mid = "white",
                       high = "tomato",    midpoint = 0) +
  labs(title = "C. Standardized Residuals",
       subtitle = "Class × Survived",
       x = "Survived", y = "Class") +
  theme_minimal(base_size = 11)

# Panel D: Partial vs zero-order (hp vs mpg)
resid_mpg <- residuals(lm(mpg ~ wt, data = mtcars))
resid_hp  <- residuals(lm(hp  ~ wt, data = mtcars))

p4 <- ggplot(data.frame(res_hp  = resid_hp,
                         res_mpg = resid_mpg),
             aes(x = res_hp, y = res_mpg)) +
  geom_point(color = "seagreen", size = 2.5, alpha = 0.8) +
  geom_smooth(method = "lm", color = "tomato",
              se = FALSE, linewidth = 1) +
  labs(title    = "D. Partial Correlation: mpg & hp | wt",
       subtitle = paste0("Partial r = ",
                          round(pcor.test(mtcars$mpg,
                                          mtcars$hp,
                                          mtcars$wt)$estimate,3)),
       x = "hp Residual", y = "mpg Residual") +
  theme_minimal(base_size = 11)

(p1 + p2) / (p3 + p4)
```

## Lab Discussion Questions

Answer the following in writing:

1. **Test Selection:** In Lab Task 2, both chi-square and Fisher's exact test give nearly identical p-values. Why? Does the very large sample size ($N = 2201$) affect which test is preferable?

2. **Effect Size vs. Significance:** The chi-square test for Class × Survival is highly significant ($p < 0.001$). Interpret Cramér's V = `r round(v_t, 3)`. Is this a practically important association? How would you communicate this to a non-statistician?

3. **Odds Ratio Interpretation:** Fisher's test for Sex × Survival gives an odds ratio. Interpret this in plain language: what does the odds ratio tell us about the relationship between being female and surviving?

4. **Partial Correlation:** After controlling for weight (`wt`), the correlation between `mpg` and `hp` weakens. Does this mean weight fully "explains" the mpg-hp relationship? What other variables might also act as confounders?

5. **Spurious Correlation:** Can you identify a potential spurious correlation in the `mtcars` dataset? Propose which third variable might be the confounder and describe how you would test this using partial correlation.

---

# Chapter Summary

## Summary

This chapter built a comprehensive toolkit for testing independence and measuring association across all combinations of variable types:

- **The concept of independence** — two variables are independent if knowing one provides no information about the other; this must be tested formally, not assumed.
- **Chi-square test** — the standard test for two categorical variables; based on comparing observed to expected frequencies in a contingency table.
- **Fisher's Exact Test** — the correct alternative when expected cell frequencies are small; computes exact probabilities rather than relying on asymptotic approximations.
- **Cramér's V** — standardizes chi-square to a 0–1 scale, enabling interpretation of association strength independently of sample size and table dimensions.
- **Pearson and Spearman correlation tests** — test whether linear (Pearson) or monotone (Spearman) relationships between continuous or ordinal variables are significantly different from zero.
- **Point-biserial and phi coefficients** — extend the correlation framework to binary variables, maintaining mathematical consistency with Pearson's $r$.
- **Partial correlation** — isolates the "pure" relationship between two variables after removing the influence of one or more control variables, helping distinguish real associations from spurious ones.

::: {.callout-important}
## Key Formulas to Know

**Chi-Square Statistic:**
$$\chi^2 = \sum_{i,j} \frac{(O_{ij} - E_{ij})^2}{E_{ij}}, \quad E_{ij} = \frac{R_i C_j}{N}$$

**Cramér's V:**
$$V = \sqrt{\frac{\chi^2/N}{\min(r-1,\, c-1)}}$$

**Pearson Correlation t-test:**
$$t = \frac{r\sqrt{n-2}}{\sqrt{1-r^2}} \sim t(n-2)$$

**First-Order Partial Correlation:**
$$r_{XY \cdot Z} = \frac{r_{XY} - r_{XZ} \cdot r_{YZ}}{\sqrt{(1-r_{XZ}^2)(1-r_{YZ}^2)}}$$

**Phi Coefficient (2×2):**
$$\phi = \frac{ad - bc}{\sqrt{(a+b)(c+d)(a+c)(b+d)}}$$
:::

## References

Anderson, E. (1935). The irises of the Gaspé Peninsula. *Bulletin of the American Iris Society*, *59*, 2–5.

Cohen, J. (1988). *Statistical power analysis for the behavioral sciences* (2nd ed.). Lawrence Erlbaum Associates.

Fox, J., & Weisberg, S. (2019). *An R companion to applied regression* (3rd ed.). SAGE.

Henderson, H. V., & Velleman, P. F. (1981). Building multiple regression models interactively. *Biometrics*, *37*(2), 391–411.

Kassambara, A. (2023). *rstatix: Pipe-friendly framework for basic statistical tests* (R package version 0.7.2). https://CRAN.R-project.org/package=rstatix

Kim Larsen. (2015). *ppcor: An R package for a fast calculation to semi-partial correlation coefficients* (R package version 1.1). https://CRAN.R-project.org/package=ppcor

Meyer, D., Zeileis, A., & Hornik, K. (2023). *vcd: Visualizing categorical data* (R package version 1.4-11). https://CRAN.R-project.org/package=vcd

Pedersen, T. L. (2022). *patchwork: The composer of plots* (R package version 1.1.2). https://CRAN.R-project.org/package=patchwork

R Core Team. (2024). *R: A language and environment for statistical computing*. R Foundation for Statistical Computing. https://www.R-project.org/

Shapiro, S. S., & Wilk, M. B. (1965). An analysis of variance test for normality. *Biometrika*, *52*(3/4), 591–611.

Sievert, C., & others. (2023). *ggcorrplot: Visualization of a correlation matrix using ggplot2* (R package version 0.1.4). https://CRAN.R-project.org/package=ggcorrplot

Spearman, C. (1904). "General intelligence," objectively determined and measured. *American Journal of Psychology*, *15*(2), 201–292.

Wei, T., & Simko, V. (2021). *corrplot: Visualization of a correlation matrix* (R package version 0.92). https://CRAN.R-project.org/package=corrplot

Wickham, H. (2016). *ggplot2: Elegant graphics for data analysis*. Springer.

Wickham, H., Averick, M., Bryan, J., Chang, W., McGowan, L. D., François, R., Grolemund, G., Hayes, A., Henry, L., Hester, J., Kuhn, M., Pedersen, T. L., Miller, E., Bache, S. M., Müller, K., Ooms, J., Robinson, D., Seidel, D. P., Spinu, V., … Yutani, H. (2019). Welcome to the tidyverse. *Journal of Open Source Software*, *4*(43), 1686.

Xie, Y. (2015). *Dynamic documents with R and knitr* (2nd ed.). CRC Press.

---

*End of Chapter 4. Proceed to Chapter 5: Data Sampling Techniques.*
